---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nlmixr2

<!-- badges: start -->
[![R build
status](https://github.com/nlmixr2/nlmixr2/workflows/R-CMD-check/badge.svg)](https://github.com/nlmixr2/nlmixr2/actions)
[![CodeFactor](https://www.codefactor.io/repository/github/nlmixr2/nlmixr2/badge)](https://www.codefactor.io/repository/github/nlmixr2/nlmixr2)
[![CRAN version](http://www.r-pkg.org/badges/version/nlmixr2)](https://cran.r-project.org/package=nlmixr2)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/grand-total/nlmixr2)](https://cran.r-project.org/package=nlmixr2)
[![CRAN total downloads](https://cranlogs.r-pkg.org/badges/nlmixr2)](https://cran.r-project.org/package=nlmixr2)
<!-- badges: end -->

The goal of nlmixr2 is to support easy and robust nonlinear mixed effects models in R

## Blog for more information

For more information about ongoing development, best practices, and news about
nlmixr2, please see the [nlmixr2 blog](https://blog.nlmixr2.org/).

## Installation

For all versions of R, we need to have a compiler setup to run `nlmixr2` and `rxode2`

### Windows compilation tools setup 

For Windows the compilers come from Rtools. For R version 4.2 and
above you need to have Rtools42, for R-4.0.x, and R-4.1.x you need
Rtools40. Download and the install from
https://cran.r-project.org/bin/windows/Rtools/

### Mac compilation tools setup

To setup the mac compilers, simply

1. Install Xcode from app store

2. Install gfortran:

   a. Download and install from https://mac.r-project.org/tools/
   
   b. Add gfortran directory to the path with: `export PATH=$PATH:/usr/local/gfortran/bin`

## R package installation 

Installation nlmixr2 itself is easiest in R-4.2.x because no further compilation is required and all supporting packages are available. From R, run:

``` r
install.packages("nlmixr2",dependencies = TRUE)
```

For R-4.0.x and R-4.1.x, the `symengine` package will need to be
downgraded to run in those earlier `R` versions.  This can be done by:

``` r
# install.packages("remotes")
remotes::install_version("symengine", version = "0.1.6")
```

followed by:

``` r
install.packages("nlmixr2",dependencies = TRUE)
```

## Checking installation

You can check that your installation is likely setup correctly with the following command after installing the `nlmixr2` package:

```{r check-install, eval=FALSE}
nlmixr2::nlmixr2CheckInstall()
```

## Development version installation

Once the compilers are setup and a compatible version of `symengine`
is installed, you can install the development version of nlmixr2 and
its nlmixr2-family dependencies either by using the r-universe or by
installing manually.

### Install using the R universe

For many people this is the fastest way to install the development
version of `nlmixr2` since it provides binaries for mac, windows for
the latest and last version of R (no need to wait for a compile).

```r
install.packages(c("dparser", "nlmixr2data", "lotri", "rxode2ll",
                   "rxode2parse", "rxode2random", "rxode2et",
                   "rxode2", "nlmixr2est", "nlmixr2extra", "nlmixr2plot",
                   "nlmixr2"),
                 repos = c('https://nlmixr2.r-universe.dev',
                           'https://cloud.r-project.org'))
```
Support packages from the R universe can also be installed for the packages in the `nlmixr2` domain:

```r
install.packages(c("xpose.nlmixr2", # Additional goodness of fit plots
                                    # baesd on xpose
                   "nlmixr2targets", # Simplify work with the
                                     # `targets` package
                   "babelmixr2", # Convert/run from nlmixr2-based
                                 # models to NONMEM, Monolix, and
                                 # initialize models with PKNCA
                   "nonmem2rx", # Convert from NONMEM to
                                # rxode2/nlmixr2-based models
                   "nlmixr2lib", # a model library and model
                                 # modification functions that
                                 # complement model piping
                   "nlmixr2rpt" # Automated Microsoft Word and
                                # PowerPoint reporting for nlmixr2
                   ),
                 repos = c('https://nlmixr2.r-universe.dev',
                           'https://cloud.r-project.org'))

# Some additional packages outside of the `nlmixr2.r-univers.dev`
# install.packages("remotes")
remotes::install_github("ggPMXdevelopment/ggPMX") # Goodness of fit plots
remotes::install_github("RichardHooijmaijers/shinyMixR") # Shiny run manager (like Piranha)
```

### Install using `remotes`

This is sure to give the latest development version

``` r
# install.packages("remotes")
remotes::install_github("nlmixr2/dparser-R")
remotes::install_github("nlmixr2/nlmixr2data")
remotes::install_github("nlmixr2/lotri")
remotes::install_github("nlmixr2/rxode2ll")
remotes::install_github("nlmixr2/rxode2parse")
remotes::install_github("nlmixr2/rxode2random")
remotes::install_github("nlmixr2/rxode2et")
remotes::install_github("nlmixr2/rxode2")
remotes::install_github("nlmixr2/nlmixr2est")
remotes::install_github("nlmixr2/nlmixr2extra")
remotes::install_github("nlmixr2/nlmixr2plot")
remotes::install_github("nlmixr2/nlmixr2")
```

Optional supporting packages can be installed like so:

``` r
# install.packages("remotes")
remotes::install_github("ggPMXdevelopment/ggPMX") # Goodness of fit
                                                  # plots
remotes::install_github("nlmixr2/xpose.nlmixr2") # Additional goodness
                                                 # of fit plots
remotes::install_github("RichardHooijmaijers/shinyMixR") # Shiny run
                                                         # manager
                                                         # (like
                                                         # Piranha)
remotes::install_github("nlmixr2/nlmixr2targets") # Simplify work with
                                                  # the `targets`
                                                  # package
remotes::install_github("nlmixr2/babelmixr2") # Convert/run from
                                              # nlmixr2-based models
                                              # to NONMEM, Monolix,
                                              # and initialize models
                                              # with PKNCA
remotes::install_github("nlmixr2/nonmem2rx") # Convert from NONMEM to
                                             # rxode2/nlmixr2-based
                                             # models
remotes::install_github("nlmixr2/nlmixr2lib") # A library of models
                                              # and model modification
                                              # functions
remotes::install_github("nlmixr2/nlmixr2rpt") # Automated Microsoft
                                              # Word and PowerPoint
                                              # reporting for nlmixr2
```

### Refreshing the installation with the latest CRAN version

If you have difficulties due to errors while compiling models, it may
be useful to reinstall all of nlmixr2 and its dependencies.  For
development versions, please use the `remotes::install_github()` or
the `install.package()` with the `r-universe` above.  For the stable version,
please use the following command:

``` r
install.packages(c("dparser", "lotri", "rxode2ll", "rxode2parse",
                   "rxode2random", "rxode2et", "rxode2",
                   "nlmixr2data", "nlmixr2est", "nlmixr2extra",
                   "nlmixr2plot", "nlmixr2"))
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example, message=FALSE}
library(nlmixr2)

## The basic model consiss of an ini block that has initial estimates
one.compartment <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv <- log(31.5); label("V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  # and a model block with the error sppecification and model specification
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    d/dt(depot) = -ka * depot
    d/dt(center) = ka * depot - cl / v * center
    cp = center / v
    cp ~ add(add.sd)
  })
}

## The fit is performed by the function nlmixr/nlmix2 specifying the model, data and estimate
fit <- nlmixr2(one.compartment, theo_sd,  est="saem", saemControl(print=0))
print(fit)
```

# Plotting outputs

## Base R Graphics
You can use base plots with the fit and it will produce a standard set of goodness of fit plots:

```{r base-r-plot, eval=FALSE}
pdf(file="myplots.pdf")
plot(fit)
dev.off()
```


## xpose.nlmixr2

The {xpose.nlmixr2} package extends xpose support for nlmixr2. You simply need to convert the fit results into an xpose database: 

```{r xpose-plot-1, eval=FALSE}
library(xpose.nlmixr2)
xpdb = xpose_data_nlmixr(fit)
```

Then you can use any of the xpose functions for generating goodness of fit plots:

```{r xpose-plot-2, eval=FALSE}
library(xpose)
plt <- dv_vs_ipred(xpdb)
```


## ggPMX

Another option is to use the ggPMX package. You first creat a ggPMX controller object from the nlmixr fit object. Then that controller object can be used to generate figures:

```{r ggpmx-plot, eval=FALSE}
library(ggPMX)
ctr = pmx_nlmixr(fit)
pmx_plot_dv_ipred(ctr)
```
