---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nlmixr2

<!-- badges: start -->
<!-- badges: end -->

The goal of nlmixr2 is to support easy and robust nonlinear mixed effects models in R

## Installation

For all versions of R, we need to have a compiler setup to run `nlmixr2` and `rxode2`

### Windows compilation tools setup 

For Windows the compilers come from Rtools. For R version 4.2 and
above you need to have Rtools42, for R-4.0.x, and R-4.1.x you need
Rtools40. Download and the install from
https://cran.r-project.org/bin/windows/Rtools/

### Mac compilation tools setup

To setup the mac compilers, simply

1. Install Xcode from app store

2. Install gfortran:

   a. Download and install from https://mac.r-project.org/tools/
   
   b. Add gfortran directory to the path with: `export PATH=$PATH:/usr/local/gfortran/bin`

# R package installation 

Installation nlmixr2 itself is easiest in R-4.2.x because no further compilation is required and all supporting packages are available. From R, run:

``` r
install.packages("nlmixr2",dependencies = TRUE)
```

For R-4.0.x and R-4.1.x, the crucial package symengine is currently not on CRAN and will have to be installed from MRAN first by running:

``` r
install.packages("symengine", repos="https://cran.microsoft.com/snapshot/2022-01-01/")
```

followed by:

``` r
install.packages("nlmixr2",dependencies = TRUE)
```

## Development version installation

Once the compilers are setup and a compatible version of `symengine`
is installed, you can install the development version of nlmixr2 and
its nlmixr2-family dependencies like so:

``` r
remotes::install_github("nlmixr2/nlmixr2data")
remotes::install_github("nlmixr2/lotri")
remotes::install_github("nlmixr2/rxode2")
remotes::install_github("nlmixr2/nlmixr2est")
remotes::install_github("nlmixr2/nlmixr2extra")
remotes::install_github("nlmixr2/nlmixr2plot")
remotes::install_github("nlmixr2/nlmixr2")
```

Optional supporting packages can be installed like so:

``` r
remotes::install_github("ggPMXdevelopment/ggPMX") # Goodness of fit plots
remotes::install_github("nlmixr2/xpose.nlmixr2") # Additional goodness of fit plots
remotes::install_github("RichardHooijmaijers/shinyMixR") # Shiny run manager (like Piranha)
remotes::install_github("nlmixr2/nlmixr2targets") # Simplify work with the `targets` package
```

If you have difficulties due to errors while compiling models, it may
be useful to reinstall all of nlmixr2 and its dependencies.  For
development versions, please use the `remotes::install_github()`
commands above.  For the stable version, please use the following
command:

``` r
install.packages(c("nlmixr2", "nlmixr2est", "rxode2", "nlmixr2plot", "nlmixr2data", "lotri", "nlmixr2extra"))
```


## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(nlmixr2)

## The basic model consiss of an ini block that has initial estimates
one.compartment <- function() {
  ini({
    tka <- 0.45 # Log Ka
    tcl <- 1 # Log Cl
    tv <- 3.45    # Log V
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v ~ 0.1
    add.sd <- 0.7
  })
  # and a model block with the error sppecification and model specification
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v <- exp(tv + eta.v)
    d/dt(depot) = -ka * depot
    d/dt(center) = ka * depot - cl / v * center
    cp = center / v
    cp ~ add(add.sd)
  })
}

## The fit is performed by the function nlmixr/nlmix2 specifying the model, data and estimate
fit <- nlmixr2(one.compartment, theo_sd,  est="saem", saemControl(print=0))
print(fit)
```

# Default plots

```{r}
plot(fit)
```
