---
title: "Delay differential equations (DDEs) in nlmixr2"
output:
  rmarkdown::html_vignette:
    self_contained: no
vignette: >
  %\VignetteIndexEntry{Delay differential equations (DDEs) in nlmixr2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
  )
library(nlmixr2save)
options(nlmixr2save.dir = "cache",
        nlmixr2save.prefix = "delays-",
        nlmixr2save.check = FALSE)
```

```{r setup}
library(nlmixr2)
library(ggplot2)
```

Most PKPD models say that the rate of change of a system depends on where
the system is *now*:

$$\frac{d}{dt}x(t) = f(t, x(t)), \qquad x(0) = x_0$$

Biology often disagrees. A cell that is dying today was marked for death by a
drug concentration it saw several days ago; a platelet entering the circulation
today was produced a week ago. In these systems the rate of change depends on
where the system *was*, which is a **delay differential equation** (DDE):

$$\frac{d}{dt}x(t) = f\big(t,\, x(t),\, x(t - T)\big), \qquad x(t) = x_0(t) \;\;\text{for}\; -T \le t \le 0$$

Koch, Krzyzanski, Pérez-Ruixo and Schropp[^koch] give a tutorial on these models
and the two features that distinguish them from ODEs:

1. the mechanism depends on the delayed state $x(t-T)$, so there is an
   **explicit delay parameter** $T$ that can be estimated; and
2. a DDE needs a **past** $x_0(t)$ on $[-T, 0]$ — a whole function, not just an
   initial value — because the system starts by reading history that predates
   $t = 0$.

[^koch]: Koch G, Krzyzanski W, Pérez-Ruixo JJ, Schropp J. Modeling of delays in
    PKPD: classical approaches and a tutorial for delay differential equations.
    *J Pharmacokinet Pharmacodyn*. 2014;41(4):291–318.
    doi:[10.1007/s10928-014-9368-y](https://doi.org/10.1007/s10928-014-9368-y)

`rxode2` (and therefore `nlmixr2`) expresses both directly:

| Concept | Syntax |
|---|---|
| the delayed state $x(t-T)$ | `delay(x, T)` |
| a non-constant past $x_0(t)$ | `past(x, T) <- expr` |

Without a `past()` statement the pre-history is the state's initial condition,
held constant — which is what most models want. `past()` is for the models that
need more, and we get to one below.

This article works two of Koch's examples end to end: we **simulate** each model
from the paper's own published parameter values, then **estimate** it back —
using the same kind of estimation the original article used for each. Example 2
was an individual fit in Berkeley Madonna, so we use a derivative-free optimizer
(`est="bobyqa"`); Example 6 was a population fit in Monolix, so we use
`est="ifoceif"` (FOCEi with a mu-referenced IRLS inner problem and the analytic
Almquist outer gradient). In both, the delay is recovered from the data.

# Why bother? Delays versus transit compartments

The usual way to put a delay into an ODE model is a chain of transit
compartments. That is a genuinely different model: a transit chain spreads the
delay out (each molecule takes a *random*, gamma-distributed time to traverse
it), whereas a lifespan model says every cell lives *exactly* $T$ days. When the
underlying biology really is a lifespan, the DDE is both more faithful and much
smaller — two equations instead of $n$.

# Example 2: lifespan-based tumour growth

Simeoni's xenograft model routes drug-damaged cells through a chain of
transit compartments $d_1 \ldots d_n$ before they die. Koch's Example 2
rewrites it so that damaged cells instead die after an *exact* lifespan $T_d$.
The whole chain collapses into one equation whose outflow is its own inflow,
delayed:

$$\frac{d}{dt}p(t) = g\big(k_0,k_1,p(t),d(t)\big) - k_{pot}\,c(t)\,p(t), \qquad p(0)=w_0$$
$$\frac{d}{dt}d(t) = \underbrace{k_{pot}\,c(t)\,p(t)}_{\text{damaged now}} - \underbrace{k_{pot}\,c(t-T_d)\,p(t-T_d)}_{\text{damaged } T_d \text{ ago, now dead}}, \qquad d(0)=0$$

with total tumour weight $w(t) = p(t) + d(t)$ and growth

$$g(k_0,k_1,p,d) = \frac{2k_0k_1p^2}{(k_1 + 2k_0p)(p+d)}$$

Here the past is *constant and zero*: no drug is given before the tumour is
inoculated, so $k_{pot}c(t)p(t) = 0$ for $-T_d \le t \le 0$. `delay()` alone
covers it — the default constant pre-history is exactly right.

The parameters below are the paper's own Berkeley Madonna values (Appendix 2):
a 100 mg dose daily on days 12–16.

```{r ex2-model}
ex2 <- rxode2({
  ka <- 103.96 * 24; ke <- 0.1052 * 24; V <- 2.7882
  l0 <- 0.195; l1 <- 0.245; w0 <- 0.010
  kpot <- 0.007; tau <- 3.61
  d/dt(x1) <- -ka * x1                      # depot
  d/dt(x2) <- ka * x1 - ke * x2             # central
  c    <- x2 / V
  cdel <- delay(x2, tau) / V                # concentration tau days ago
  ## untreated control tumour, for reference
  d/dt(x3) <- (2 * l0 * l1 * x3) / (l1 + 2 * l0 * x3)
  ## proliferating cells
  d/dt(x4) <- (2 * l0 * l1 * x4^2) / ((l1 + 2 * l0 * x4) * (x4 + x5)) - kpot * c * x4
  ## apoptotic cells: damaged now, minus damaged exactly tau days ago
  d/dt(x5) <- kpot * c * x4 - kpot * cdel * delay(x4, tau)
  x3(0) <- w0; x4(0) <- w0; x5(0) <- 0
  w <- x4 + x5                              # total tumour weight
})
```

A DDE has to interpolate its own history, so the solver must retain it: pass
`dense=TRUE` (and a suitable method) when solving by hand.

```{r ex2-sim}
ev2 <- et(seq(0, 30, by = 0.25))
for (td in 12:16) ev2 <- et(ev2, amt = 300, cmt = "x1", time = td)

s2 <- rxSolve(ex2, ev2, method = "dop853", dense = TRUE, atol = 1e-10, rtol = 1e-10)

ggplot(s2, aes(time)) +
  geom_line(aes(y = x3, colour = "untreated control")) +
  geom_line(aes(y = w, colour = "treated")) +
  annotate("rect", xmin = 12, xmax = 16, ymin = -Inf, ymax = Inf, alpha = 0.1) +
  labs(x = "time (days)", y = "tumour weight (g)", colour = NULL,
       title = "Koch Example 2: lifespan-based tumour growth",
       subtitle = "shaded band = dosing days 12-16") +
  theme_bw() + theme(legend.position = "top")
```

## Simulating a population

To show the delay can be *recovered*, we simulate 24 mice with between-subject
variability on the drug potency `kpot` and on the apoptotic lifespan `tau`
itself, and observe tumour weight with additive error.

```{r ex2-data}
sim2 <- rxode2({
  ka <- 103.96 * 24; ke <- 0.1052 * 24; V <- 2.7882
  l0 <- 0.195; l1 <- 0.245; w0 <- 0.010
  kpot <- 0.007 * exp(eta.kpot)
  tau  <- 3.61 * exp(eta.tau)
  d/dt(x1) <- -ka * x1
  d/dt(x2) <- ka * x1 - ke * x2
  c    <- x2 / V
  cdel <- delay(x2, tau) / V
  d/dt(x4) <- (2 * l0 * l1 * x4^2) / ((l1 + 2 * l0 * x4) * (x4 + x5)) - kpot * c * x4
  d/dt(x5) <- kpot * c * x4 - kpot * cdel * delay(x4, tau)
  x4(0) <- w0; x5(0) <- 0
  w <- x4 + x5
})

set.seed(101)
obst <- c(0, 3, 6, 9, 12, 14, 16, 18, 20, 22, 24, 26, 28, 30)
N <- 16
dose <- 300     # a strong drug effect makes the delayed apoptotic removal --
ev <- et(obst)  # and hence the lifespan -- observable in the tumour weight
for (td in 12:16) ev <- et(ev, amt = dose, cmt = "x1", time = td)

rows <- list()
for (i in 1:N) {
  p <- c(eta.kpot = rnorm(1, 0, sqrt(0.04)), eta.tau = rnorm(1, 0, sqrt(0.02)))
  s <- rxSolve(sim2, params = p, ev, method = "dop853", dense = TRUE,
               returnType = "data.frame", atol = 1e-8, rtol = 1e-8)
  s <- s[s$time %in% obst, ]
  dv <- s$w + rnorm(nrow(s), 0, 0.02); dv[dv <= 0] <- 1e-3
  rows[[i]] <- data.frame(ID = i, TIME = s$time, DV = dv, AMT = 0, EVID = 0, CMT = "x4")
}
obs <- do.call(rbind, rows)
dos <- do.call(rbind, lapply(1:N, function(i)
  data.frame(ID = i, TIME = 12:16, DV = NA_real_, AMT = dose, EVID = 1, CMT = "x1")))
dat2 <- rbind(obs, dos)
dat2 <- dat2[order(dat2$ID, dat2$TIME, -dat2$EVID), ]
head(dat2)
```

## Estimating it back

Koch et al. fit this example in Berkeley Madonna as an **individual** model —
no random effects — and estimated the lifespan `Td` directly from the data
(their Table 1). We do the same: a single-subject / naive-pooled fit estimating
the drug potency `kpot` and the apoptotic **lifespan** `tau` (the delay).

The choice of *optimizer* matters here, and it is worth understanding why. The
delay parameter sits inside a product of two delayed states,
`delay(x2, tau) * delay(x4, tau)`, and the PK is stiff. A parameter like that is
awkward for a *gradient-based* optimizer: the model's exact derivative with
respect to the delay time is expensive, and a finite-difference gradient is
noisy because moving `tau` slides the interpolated history. So we use
`est="bobyqa"` — Powell's **derivative-free** bounded optimizer, which only ever
*evaluates* the objective and never differentiates it. That is exactly the kind
of optimizer Berkeley Madonna uses, and it walks straight to the minimum.

```{r ex2-est-model}
mod2 <- function() {
  ini({
    tkpot <- c(log(0.002), log(0.006), log(0.02))   # bounded; truth 0.007
    ttau  <- c(log(1.5),   log(3.0),   log(6.0))    # bounded; Td, truth 3.61
    add.err <- c(0.001, 0.02, 0.2)
  })
  model({
    ka <- 103.96 * 24; ke <- 0.1052 * 24; V <- 2.7882
    l0 <- 0.195; l1 <- 0.245; w0 <- 0.010
    kpot <- exp(tkpot)
    tau  <- exp(ttau)
    d/dt(x1) <- -ka * x1
    d/dt(x2) <- ka * x1 - ke * x2
    c    <- x2 / V
    cdel <- delay(x2, tau) / V
    d/dt(x4) <- (2 * l0 * l1 * x4^2) / ((l1 + 2 * l0 * x4) * (x4 + x5)) - kpot * c * x4
    d/dt(x5) <- kpot * c * x4 - kpot * cdel * delay(x4, tau)
    x4(0) <- w0; x5(0) <- 0
    w <- x4 + x5
    w ~ add(add.err)
  })
}
```

`optExpression=FALSE` turns off common-subexpression optimization: a `delay()`
term has to be regenerated together with the `d/dt()` of the state it delays,
which defeats the optimizer's model-splitting on a DDE and makes assembly slow.
Turning it off skips that; the model solves fine without it.

```{r ex2-fit}
fit2 := nlmixr2(mod2, dat2, est = "bobyqa",
                control = bobyqaControl(optExpression = FALSE))
fit2
```

```{r ex2-recovery}
truth2 <- c(kpot = 0.007, tau = 3.61)
est2 <- exp(fixef(fit2)[c("tkpot", "ttau")])
data.frame(parameter = names(truth2),
           truth = truth2,
           estimate = round(est2, 4),
           ratio = round(est2 / truth2, 3),
           row.names = NULL)
```

The apoptotic lifespan `tau` is recovered close to its true 3.61 days — the
delay is estimated directly from the tumour-weight data, exactly as in the
original article.

# Example 6: a model whose past is not constant

Example 2 got away with the default pre-history because nothing happened before
$t=0$. Koch's Example 6 does not.

In collagen-induced arthritic mice, arthritis has an **induction phase** with no
measurements, followed by a **development phase** where data collection starts.
The cytokine GM-CSF is already being over-produced during induction, so at
$t = 0$ the system is mid-flight: its history is a growing exponential, not a
constant. The model (a total arthritic score $R_1$, and a strongly delayed
ankylosis score $R_2$) reads

$$\frac{d}{dt}G(t) = k_3 - \frac{k_1}{k_2}\big(1 - e^{-k_2 t}\big)G(t) - \frac{E_{max}c(t)}{EC_{50}+c(t)}G(t), \qquad \boxed{G(t) = a e^{bt} \;\text{for}\; -T \le t \le 0}$$
$$\frac{d}{dt}I(t) = k_4 G(t) - k_4 G(t-T), \qquad I(0)=I_0$$
$$\frac{d}{dt}D(t) = k_4 G(t-T) - k_5 D(t), \qquad D(0)=0$$
$$R_1(t) = I(t) + D(t) \quad\text{(TAS)}, \qquad R_2(t) = D(t) \quad\text{(AKS)}$$

That boxed line is the whole point. $D$ is driven by $G(t-T)$ with $T \approx 13$
days, so for the first two weeks the ankylosis score is driven **entirely by
cytokine history that predates the study**. Get the past wrong and the entire
induction/early-response phase is wrong.

The paper implements this in MONOLIX, where the delayed state's initial
condition is written as an expression in `t` (`y1_0 = a*exp(b*t)`). `rxode2`
says the same thing with `past()`:

```{r ex6-model}
ex6 <- rxode2({
  k  <- 0.184; V <- 0.039
  k1 <- 0.535; k2 <- 0.352; k3 <- 5
  k4 <- 0.122; k5 <- 0.126
  Emax <- 1; EC50 <- 44.3
  T  <- 12.8; I0 <- 2.42
  a  <- 1; b <- 0.5
  d/dt(Ac) <- -k * Ac
  c   <- Ac / V
  eff <- (Emax * c) / (EC50 + c)
  G(0) <- a                     # continuity: a*exp(b*0) = a
  d/dt(G) <- k3 - eff * G - (k1 / k2) * (1 - exp(-k2 * t)) * G
  past(G, T) <- a * exp(b * t)  # <- the induction-phase history, on [-T, 0]
  I(0) <- I0
  d/dt(I) <- k4 * G - k4 * delay(G, T)
  D(0) <- 0
  d/dt(D) <- k4 * delay(G, T) - k5 * D
  TAS <- I + D
  AKS <- D
})
```

## Checking the past against the paper

Koch derives the model's own closed form on $[0, T]$ (their Eqs. 85–87): while
$t \le T$, the delayed cytokine is just the history evaluated at $t-T$, i.e.
$G(t-T) = a\,e^{b(t-T)}$. That is an exact statement we can test.

```{r ex6-check}
probe <- rxode2({
  T <- 12.8; a <- 1; b <- 0.5
  k3 <- 5; k1 <- 0.535; k2 <- 0.352
  G(0) <- a
  d/dt(G) <- k3 - (k1 / k2) * (1 - exp(-k2 * t)) * G
  past(G, T) <- a * exp(b * t)
  gdel <- delay(G, T)
})
tt <- seq(0, 12.8, by = 0.1)
sp <- rxSolve(probe, et(tt), method = "dop853", dense = TRUE, atol = 1e-10, rtol = 1e-10)
max(abs(sp$gdel - 1 * exp(0.5 * (tt - 12.8))))   # vs the paper's Eq (86)/(87)
```

Zero — `past()` reproduces the published closed form exactly.

It is worth seeing what the *default* constant pre-history would have done here,
since that is the trap this syntax exists to avoid:

```{r ex6-wrong}
ex6const <- rxode2({
  k <- 0.184; V <- 0.039; k1 <- 0.535; k2 <- 0.352; k3 <- 5
  k4 <- 0.122; k5 <- 0.126; Emax <- 1; EC50 <- 44.3
  T <- 12.8; I0 <- 2.42; a <- 1
  d/dt(Ac) <- -k * Ac
  c   <- Ac / V
  eff <- (Emax * c) / (EC50 + c)
  G(0) <- a
  d/dt(G) <- k3 - eff * G - (k1 / k2) * (1 - exp(-k2 * t)) * G
  I(0) <- I0
  d/dt(I) <- k4 * G - k4 * delay(G, T)
  D(0) <- 0
  d/dt(D) <- k4 * delay(G, T) - k5 * D
  AKS <- D
})

ev6 <- et(seq(0, 30, by = 0.25))
for (td in c(1, 8, 15)) ev6 <- et(ev6, amt = 1, cmt = "Ac", time = td)
s6 <- rxSolve(ex6, ev6, method = "dop853", dense = TRUE, atol = 1e-10, rtol = 1e-10)
s6c <- rxSolve(ex6const, ev6, method = "dop853", dense = TRUE, atol = 1e-10, rtol = 1e-10)

cmp <- rbind(
  data.frame(time = s6$time,  AKS = s6$AKS,  past = "past(G,T) <- a*exp(b*t)"),
  data.frame(time = s6c$time, AKS = s6c$AKS, past = "default constant past"))

ggplot(cmp, aes(time, AKS, colour = past)) +
  geom_line() +
  geom_vline(xintercept = 12.8, linetype = 3) +
  labs(x = "time (days)", y = "ankylosis score D(t)", colour = NULL,
       title = "Koch Example 6: the past is part of the model",
       subtitle = "dotted line = the delay T; before it, D(t) is driven purely by history") +
  theme_bw() + theme(legend.position = "top")
```

## Simulating and estimating

We simulate a small population with three endpoints — PK (proportional error),
TAS and AKS (additive) — dosing at days 1, 8 and 15, using the paper's Table 5
population values.

```{r ex6-data}
sim6 <- rxode2({
  k  <- 0.184 * exp(eta.k)
  V  <- 0.039
  k1 <- 0.535; k2 <- 0.352; k3 <- 5
  k4 <- 0.122 * exp(eta.k4)
  k5 <- 0.126 * exp(eta.k5)
  Emax <- 1; EC50 <- 44.3
  T  <- 12.8 * exp(eta.T)
  I0 <- 2.42 * exp(eta.I0)
  a  <- 1; b <- 0.5
  d/dt(Ac) <- -k * Ac
  c   <- Ac / V
  eff <- (Emax * c) / (EC50 + c)
  G(0) <- a
  d/dt(G) <- k3 - eff * G - (k1 / k2) * (1 - exp(-k2 * t)) * G
  past(G, T) <- a * exp(b * t)
  I(0) <- I0
  d/dt(I) <- k4 * G - k4 * delay(G, T)
  D(0) <- 0
  d/dt(D) <- k4 * delay(G, T) - k5 * D
  TAS <- I + D
  AKS <- D
})

set.seed(202)
pkT <- c(1.0033, 3, 5, 7, 8.0033, 10, 15.0033, 18, 21)
pdT <- c(0, 1, 3, 5, 7, 8, 10, 12, 15, 18, 21, 25, 28)
N <- 12
ev <- et(sort(unique(c(pkT, pdT))))
for (td in c(1, 8, 15)) ev <- et(ev, amt = 1, cmt = "Ac", time = td)

rows <- list()
for (i in 1:N) {
  p <- c(eta.k = rnorm(1, 0, sqrt(0.212)), eta.k4 = rnorm(1, 0, sqrt(0.279)),
         eta.k5 = rnorm(1, 0, sqrt(0.3)),  eta.T = rnorm(1, 0, sqrt(0.0782)),
         eta.I0 = rnorm(1, 0, sqrt(0.562)))
  s <- rxSolve(sim6, params = p, ev, method = "dop853", dense = TRUE,
               returnType = "data.frame", atol = 1e-8, rtol = 1e-8)
  spk <- s[s$time %in% pkT, ]
  spd <- s[s$time %in% pdT, ]
  cdv <- spk$c * (1 + rnorm(nrow(spk), 0, 0.382)); cdv[cdv <= 0] <- 1e-3
  rows[[i]] <- rbind(
    data.frame(ID = i, TIME = spk$time, DV = cdv, AMT = 0, EVID = 0, CMT = "c"),
    data.frame(ID = i, TIME = spd$time, DV = spd$TAS + rnorm(nrow(spd), 0, 1.49),
               AMT = 0, EVID = 0, CMT = "TAS"),
    data.frame(ID = i, TIME = spd$time, DV = spd$AKS + rnorm(nrow(spd), 0, 0.420),
               AMT = 0, EVID = 0, CMT = "AKS"))
}
obs6 <- do.call(rbind, rows)
dos6 <- do.call(rbind, lapply(1:N, function(i)
  data.frame(ID = i, TIME = c(1, 8, 15), DV = NA_real_, AMT = 1, EVID = 1, CMT = "Ac")))
dat6 <- rbind(obs6, dos6)
dat6 <- dat6[order(dat6$ID, dat6$TIME, -dat6$EVID), ]
```

We estimate the four parameters that carry the DDE structure — the ankylosis
delay `T`, the inflammation and destruction rates `k4` and `k5`, and, most
interestingly, `b` — with between-subject variability on the delay, and the
three residual errors, fixing the PK and cytokine-turnover constants at the
published values.

`b` deserves emphasis. It is the growth rate of the induction-phase history
`a*exp(b*t)`, so it appears **only before $t=0$** — the model never evaluates it
at any observed time point directly. It is identified purely through the
history's downstream effect on the delayed ankylosis score. Estimating it is a
thing you simply cannot express without a non-constant `past()`.

```{r ex6-est-model}
mod6 <- function() {
  ini({
    tk4 <- log(0.15)
    tk5 <- log(0.10)
    tT  <- log(11)         # delay, true 12.8
    tb  <- log(0.40)       # history growth rate, true 0.5
    eta.T ~ 0.0782
    prop.pk <- 0.382
    add.tas <- 1.49
    add.aks <- 0.420
  })
  model({
    k  <- 0.184; V <- 0.039
    k1 <- 0.535; k2 <- 0.352; k3 <- 5
    k4 <- exp(tk4)
    k5 <- exp(tk5)
    Emax <- 1; EC50 <- 44.3
    T  <- exp(tT + eta.T)
    I0 <- 2.42
    a  <- 1
    b  <- exp(tb)
    d/dt(Ac) <- -k * Ac
    c   <- Ac / V
    eff <- (Emax * c) / (EC50 + c)
    G(0) <- a
    d/dt(G) <- k3 - eff * G - (k1 / k2) * (1 - exp(-k2 * t)) * G
    past(G, T) <- a * exp(b * t)
    I(0) <- I0
    d/dt(I) <- k4 * G - k4 * delay(G, T)
    D(0) <- 0
    d/dt(D) <- k4 * delay(G, T) - k5 * D
    TAS <- I + D
    AKS <- D
    c   ~ prop(prop.pk)
    TAS ~ add(add.tas)
    AKS ~ add(add.aks)
  })
}
```

```{r ex6-fit}
fit6 := nlmixr2(mod6, dat6, est = "ifoceif",
                control = foceiControl(print = 0, covMethod = "", optExpression = FALSE))
fit6
```

```{r ex6-recovery}
truth6 <- c(k4 = 0.122, k5 = 0.126, T = 12.8, b = 0.5)
est6 <- exp(fixef(fit6)[c("tk4", "tk5", "tT", "tb")])
data.frame(parameter = names(truth6),
           truth = truth6,
           estimate = round(est6, 4),
           ratio = round(est6 / truth6, 3),
           row.names = NULL)
```

The delay `T` comes back close to its true 12.8 days, and — remarkably — so does
the pre-history growth rate `b`, even though `b` describes cytokine dynamics that
happened before the study began and are never observed directly. That a
parameter of the past is estimable at all is the whole point of a non-constant
`past()`.

# Practical notes

- **Keep the history.** A DDE interpolates its own past, so the solver has to
  record it. `nlmixr2` arranges this for you when it sees a `delay()`; when you
  call `rxSolve()` yourself, pass `dense=TRUE` and a dense-output method such as
  `method="dop853"`.
- **`past()` is only for a non-constant history.** The default pre-history is the
  initial condition held constant, which is what Example 2 (and most lifespan
  models) want. Reach for `past()` when the system was already running before
  $t=0$, as in Example 6.
- **The delay is a parameter.** $T$ is estimated like anything else and can carry
  a random effect; that is the main practical advantage over a transit chain,
  where the delay is implied by a rate constant and a fixed number of
  compartments.
- **Choose the optimizer to match the delay.** A delay time is a difficult
  parameter for a gradient-based optimizer: its exact derivative is expensive,
  and a finite-difference gradient is noisy because perturbing the delay slides
  the interpolated history. When a delay is well constrained by the data
  (Example 6's ankylosis score), the gradient methods still succeed. When it is
  not, or when it sits inside a stiff system and a product of delayed states
  (Example 2), a **derivative-free** optimizer such as `est="bobyqa"` — which
  only evaluates the objective — is far more robust. Bounding the parameters
  keeps such a fit from wandering into regions where the stiff solve blows up.

# Reproducing this article

The two fits here are expensive, so they are cached under `vignettes/cache/`
and reused by the website build. To refit from scratch:

```sh
cd vignettes && Rscript precompute.R --clean
```
