---
title: "Contributing a long-running example (precompute + cache)"
output:
  rmarkdown::html_vignette:
    self_contained: no
vignette: >
  %\VignetteIndexEntry{Contributing a long-running example (precompute + cache)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE            # this is documentation; nothing here is run
)
```

This article is for **contributors**. It explains how to add a pkgdown article
whose model fits are too slow to run on every website build — a full FOCEi or
SAEM estimation, a PBPK model, a delay differential equation — without timing
the build out.

The mechanics are a one-time setup chunk plus a single line per fit.

# The problem

The articles under `vignettes/` are `.Rbuildignore`d, so they are built **only
for the website** (`pkgdown`), never during `R CMD check` or on CRAN. That is
deliberate: it lets an article run a real estimation.

But a real estimation is slow. When several articles each re-fit a SAEM or FOCEi
model on every site build, the pkgdown runner times out (it exits 143 after its
wall-clock limit). So we cache: fit once, commit the result, and load it on every
later build.

We do this with the **`:=` operator** from the
[nlmixr2save](https://github.com/nlmixr2/nlmixr2save) package, which saves a fit
in a portable format (readable independent of the nlmixr2/rxode2 version) and
reloads it instead of refitting.

# The setup chunk

Each article that fits a model starts with:

```{r}
library(nlmixr2save)
options(nlmixr2save.dir    = "cache",         # where the cache files live
        nlmixr2save.prefix = "modelPiping-",  # a per-article file prefix
        nlmixr2save.check  = FALSE)           # trust the committed cache
```

- **`nlmixr2save.dir = "cache"`** puts the cache files under `vignettes/cache/`.
- **`nlmixr2save.prefix`** namespaces this article's files. Use the article's
  base name plus a dash (`"modelPiping-"`, `"wbc-"`, `"delays-"`, …) so files do
  not collide between articles.
- **`nlmixr2save.check = FALSE`** tells `:=` to *trust* the committed cache: if
  the cache file exists it is loaded as-is; it is regenerated only when it is
  missing (or you clear it). This keeps a committed cache stable across
  nlmixr2/rxode2 updates — the fits are re-run only when you actually want them
  re-run.

# Caching a fit with `:=`

Replace a normal assignment `fit <- nlmixr2(...)` with `fit := nlmixr2(...)`:

```{r}
fit := nlmixr2(one.compartment, theo_sd, est = "focei",
               control = list(print = 0),
               table = list(cwres = TRUE, npde = TRUE))
```

The first time this runs it fits the model and writes
`cache/modelPiping-fit.zip`; every later build loads that file and assigns `fit`
without refitting. The variable name (`fit`) plus the prefix is the cache file
name, so **each `:=` target must be unique within an article**.

A simulation or other non-fit value works the same way and is stored as an
`.rds` (a fit becomes a portable `.zip`):

```{r}
vpc := vpcSim(fit, n = 300)     # cache/modelPiping-vpc.rds
```

## One model transformation per fit

`:=` caches a single `nlmixr2()`/`nlmixr()` call. When an example builds its
model by piping, do the (cheap) model transformation first and give the (slow)
fit its own `:=` line:

```{r}
# build the model -- cheap, not cached
noEta.mod <- fit |> model(ka <- exp(tka))

# fit it -- cached
noEta := nlmixr2(noEta.mod, theo_sd, est = "focei",
                 control = list(print = 0))
```

Piping a fit through `ini()`/`model()`/`update()` returns a model (an `rxUi`),
which carries no data, so pass the data explicitly to `nlmixr2()`.

# Registering, populating, and committing the cache

**1. Register the article** in `precompute.R` — add its file name to the
`vignettes` vector so the cache-populating script renders it:

```{r}
vignettes <- c(
  "addingCovariances.Rmd",
  "delays.Rmd",
  "mymodel.Rmd",          # <- add here
  "modelPiping.Rmd",
  # ...
)
```

**2. Populate the cache.** From the `vignettes/` directory:

```sh
cd vignettes
Rscript precompute.R          # fits only what is missing from cache/
```

`precompute.R` renders each registered article in its **own fresh R subprocess**
(building many rxode2 models in one long-lived session eventually fails), which
runs the article's real code — so the `:=` lines do the fitting and write the
cache files.

**3. Commit the cache** alongside the `.Rmd`:

```sh
git add vignettes/mymodel.Rmd vignettes/cache/mymodel-*.zip
```

**4. Add the article to the pkgdown navbar** in `_pkgdown.yml`.

# Refreshing the cache

To force the fits to re-run — after changing a model, data, or package — clear
the cache and re-render. `nlmixr2saveInvalidate()` removes every cache entry
under the active prefix/dir:

```{r}
options(nlmixr2save.dir = "cache", nlmixr2save.prefix = "mymodel-")
nlmixr2saveInvalidate()       # clears cache/mymodel-*
```

or clear everything and rebuild all articles:

```sh
cd vignettes
Rscript precompute.R --clean   # empties cache/ then refits every article
```

# When a fit needs settings that only matter for the demo

A cached fit is computed once, so it is fine — and often clearer — to use
control settings that make the *demonstration* fast and robust rather than
production-grade. The DDE article, for example, skips the standard-error step
(`covMethod = ""`) and common-subexpression optimization
(`optExpression = FALSE`) because the point is parameter recovery, not
precision. Put such choices in the `.Rmd` (the source of truth) and note *why*
in the prose.

# Summary

- Articles in `vignettes/` build for the website only; they may run real fits.
- Set `nlmixr2save.dir`/`.prefix`/`.check` once, then cache every fit with
  `name := nlmixr2(...)` (a fit → `.zip`, a simulation → `.rds`).
- Do model piping in a separate, cheap step; give each fit its own unique `:=`.
- Register the article in `precompute.R`, run it, and commit `cache/<prefix>*`.
- `precompute.R --clean` (or `nlmixr2saveInvalidate()`) refreshes the cache when
  a model changes.
