---
title: "Contributing a long-running example (precompute + cache)"
output:
  rmarkdown::html_vignette:
    self_contained: no
vignette: >
  %\VignetteIndexEntry{Contributing a long-running example (precompute + cache)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE            # this is documentation; nothing here is run
)
```

This article is for **contributors**. It explains how to add a pkgdown article
whose model fits are too slow to run on every website build — a full FOCEi or
SAEM estimation, a PBPK model, a delay differential equation — without timing
the build out.

If you are writing a *user*-facing tutorial, read this once and then follow the
pattern; the mechanics are three lines of boilerplate.

# The problem

The articles under `vignettes/` are `.Rbuildignore`d, so they are built **only
for the website** (`pkgdown`), never during `R CMD check` or on CRAN. That is
deliberate: it lets an article run a real estimation.

But a real estimation is slow. When several articles each re-fit a SAEM or FOCEi
model on every site build, the pkgdown runner times out (it exits 143 after its
wall-clock limit). The larger models — the PBPK mavoglurant fit, the
myelosuppression model, the DDE examples — are minutes each.

We do **not** want to fake the fits, and we do **not** want to hand-maintain a
copy of the numbers separately from the code that produces them. So we cache.

# The pattern: `cacheFit()`

Every article that fits a model starts by sourcing the cache helper:

```{r}
source("cache.R")
```

Then each expensive estimation call is wrapped in `cacheFit(name, expr)`:

```{r}
fit <- cacheFit("wbc-fit.S", {
  nlmixr(wbc, d3, est = "saem", list(print = 0),
         table = list(cwres = TRUE, npde = TRUE))
})
```

`cacheFit()` does exactly one thing:

- if `vignettes/cache/<name>.rds` **exists**, it loads and returns that fit —
  the `expr` is never evaluated (it is an ordinary lazy argument), so the site
  build does no fitting;
- if it does **not** exist, it evaluates `expr`, saves the result to
  `cache/<name>.rds` (xz-compressed), and returns it.

The `.rds` files are committed to the repository. On the website the fit always
comes from the cache; the estimation code in the `.Rmd` is the single source of
truth for how that cache was produced.

## Naming

Give each cached fit a unique, article-prefixed name, e.g. `"wbc-fit.S"`,
`"delays-ex6-ifoceif"`, `"mavoglurant-fit.lnorm.S"`. Two articles must not share
a cache name.

# Adding a new article: the checklist

Suppose you are adding `mymodel.Rmd`.

**1. Source the cache and wrap every fit.**

```{r}
# top of mymodel.Rmd, in a setup chunk
source("cache.R")
```

```{r}
fit <- cacheFit("mymodel-mainfit", {
  nlmixr2(model, data, est = "focei")
})
```

**2. Register the article in `precompute.R`.** Add its file name to the
`vignettes` vector so the cache-populating script knows to render it:

```{r}
vignettes <- c(
  "addingCovariances.Rmd",
  "broom.Rmd",
  "delays.Rmd",
  "mymodel.Rmd",          # <- add here
  "mavoglurant.Rmd",
  # ...
)
```

**3. Populate the cache.** From the `vignettes/` directory:

```sh
cd vignettes
Rscript precompute.R          # fits only what is missing from cache/
```

`precompute.R` renders each registered article to a throwaway location. Because
rendering runs the article's real code, the `cacheFit()` calls inside it do the
fitting and write the `.rds` files. Each article is rendered in its **own fresh
R subprocess** — building many rxode2 models in one long-lived session
accumulates loaded model DLLs and eventually fails with "error building model",
so isolating each render keeps precompute reliable.

**4. Commit the cache.** Add the new `.rds` file(s) alongside your `.Rmd`:

```sh
git add vignettes/mymodel.Rmd vignettes/cache/mymodel-mainfit.rds
```

**5. Add it to the pkgdown navbar** in `_pkgdown.yml` (under `articles`) so it
appears on the site.

# Refreshing the cache

When you change a cached model or its data, the stored fit is stale. Refit
everything from scratch:

```sh
cd vignettes
Rscript precompute.R --clean   # clears cache/ first, then refits all articles
```

`--clean` calls `cacheClean()`, which deletes every `.rds` under `cache/`. A
plain `Rscript precompute.R` (no `--clean`) only fills in what is missing, which
is what you want after adding a single new article.

# Verifying the site loads from cache (does not refit)

pkgdown renders quietly, so it is easy to *think* the cache is being used when
it is silently refitting. To check, point `NLMIXR_VIGNETTE_CACHE_TRACE` at a
file before building; `cacheFit()` writes one line per call recording whether it
was a `hit` or a `miss`:

```sh
NLMIXR_VIGNETTE_CACHE_TRACE=/tmp/cache-trace.tsv Rscript -e \
  'pkgdown::build_article("wbc")'
cat /tmp/cache-trace.tsv     # every line should say "hit"
```

A `miss` during a website build means the committed `.rds` is missing or its
name does not match the `cacheFit()` call.

The cache directory itself can be redirected with `NLMIXR_VIGNETTE_CACHE`
(default `cache`), which is occasionally useful for testing.

# When a fit needs settings that only matter for the demo

A cached fit is computed once, so it is fine — and often clearer — to use
control settings that make the *demonstration* fast and robust rather than
production-grade. For example, the DDE article skips the standard-error step
(`covMethod = ""`) and common-subexpression optimization
(`optExpression = FALSE`) because the point is parameter recovery, not
precision, and those settings keep a delay-model fit tractable. Put such choices
in the `.Rmd` (the source of truth) and note *why* in the prose, so the next
person understands the fit they are looking at.

# Summary

- Articles in `vignettes/` build for the website only; they may run real fits.
- Wrap every expensive fit in `cacheFit("unique-name", { ... })`.
- Register the article in `precompute.R`, run `Rscript precompute.R`, and commit
  the resulting `cache/*.rds`.
- The website loads the committed cache and never refits; the `.Rmd` stays the
  single source of truth.
- `precompute.R --clean` refreshes everything when a model changes.
