---
title: "Mixture Models in nlmixr2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixture Models in nlmixr2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4
)
library(nlmixr2)
library(nlmixr2save)
## Every fit below is REAL and its output is whatever the current nlmixr2
## produced -- nothing in this article is transcribed by hand.  The `:=`
## operator (nlmixr2save) runs each fit once and caches it under the prefix
## below, so the site rebuilds without refitting.  The shared cache ships with
## the package in inst/cache/ and is located with system.file().
options(nlmixr2save.dir = system.file("cache", package = "nlmixr2"),
        nlmixr2save.prefix = "mixture-",
        nlmixr2save.check = FALSE)
```

## Overview

Mixture models describe populations where individuals belong to one of several
latent subpopulations, each with distinct fixed-effect parameters. The classic
pharmacokinetic example is bimodal clearance due to a polymorphic
drug-metabolizing enzyme: *extensive metabolizers* (EM) clear the drug quickly
while *poor metabolizers* (PM) clear it slowly. A standard single-population
model will fit neither group well and will produce inflated between-subject
variability for clearance.

`nlmixr2est` supports mixture models via the `mix()` function inside the model
block. The following estimation methods all support mixture models:

| Method    | Description |
|-----------|-------------|
| `focei`   | First-Order Conditional Estimation with Interaction (recommended) |
| `foce`    | First-Order Conditional Estimation |
| `fo`      | First-Order |
| `foi`     | First-Order with Interaction |
| `laplace` | Laplace approximation |
| `agq`     | Adaptive Gaussian Quadrature |
| `saem`    | Stochastic Approximation Expectation-Maximization |
| `imp`     | Importance sampling (Monte-Carlo EM) |
| `impmap`  | Importance sampling around the MAP estimate |
| `qrpem`   | Quasi-random parametric EM |
| `npag`    | Nonparametric adaptive grid |
| `npb`     | Nonparametric Bayesian (Dirichlet-process) |

Two methods **reject** a `mix()` model up front rather than fitting something
misleading: `emvi` and `fbvi` (mean-field variational inference on the
theta-sensitivity model, which has no mixture structure) stop with
`cannot have a mixture model (ie 'mix()') for the estimation routine`.
`vae` does not currently fit a mixture model either.

The gradient-based methods (`focei`, `foce`, `fo`, `foi`, `laplace`, `agq`) all
use the same *hard-assignment* strategy described below and reliably separate
mixture components with high precision. SAEM's mixture machinery is a
*soft-assignment* (EM responsibility) scheme, described in its own section,
[Mixture Models with SAEM](#mixture-models-with-saem) — though its default,
`mixProbMethod = "regress"`, hard-classifies each subject once and skips the
per-iteration responsibility step, which is what makes the default immune to
the mixing-probability collapse that scheme is otherwise prone to. Even so, on
small or weakly-separated datasets it may separate components less precisely
than the gradient-based methods. **`focei` is recommended when
precise per-subject classification matters; SAEM is a usable alternative that
may need tuning (see its section) to match it.** The EM/sampling and
nonparametric families are covered in
[Other estimation families](#other-estimation-families).

> **rxode2 requirements.** Nested transformations inside `mix()` — for example
> `mix(expit(tcl1 + eta.cl, 0.1, 200), p1, ...)` — require **rxode2 >= 5.1.3**.
> Earlier versions translate only the top-level `mix()` arguments and will
> silently mis-handle the inner `expit()`/`logit()`/`exp()` calls.
>
> Reading `mixest`/`mixnum` back out of the *fitted table* (see
> [Reading the component off the fit table](#reading-the-component-off-the-fit-table))
> additionally requires an rxode2 carrying the fix in
> [nlmixr2/rxode2#1358](https://github.com/nlmixr2/rxode2/pull/1358), which is
> newer than 5.1.7. This was a bug, not a design limit: up to that version the
> prediction model lost its `mix()` call to symengine and rxode2 no longer read
> it as a mixture, so those columns — and every model variable computed from
> `mix()` — came back as `0` for every row, with `PRED`/`IPRED` computed from
> those zeros. You do not have to check the version: a fit on an affected
> rxode2 now says so in `$runInfo`
> (`"mixture not passed to table; mixest/mixnum read 0"`). The **parameter
> estimates, `$mixNum` and `$mixList` were never affected** — the bug was
> confined to the table.

## Motivating Example: Bimodal Clearance

To illustrate the approach we simulate a 30-subject study with two distinct
clearance populations — roughly mimicking a drug whose metabolism is governed by
a CYP2D6 polymorphism.

```{r simulate, message=FALSE}
library(nlmixr2est)
library(rxode2)

set.seed(2024)

# True population parameters
tclEm   <- log(8)    # fast metabolizers (EM), CL = 8 L/hr
tclPm   <- log(0.8)  # slow metabolizers (PM), CL = 0.8 L/hr  (10× lower)
tv      <- log(30)   # volume of distribution
tka     <- log(1.2)  # absorption rate constant
sigma   <- 0.20      # proportional residual SD
omegaCl <- 0.09      # BSV on CL (variance)
omegaV  <- 0.04      # BSV on V

nEm <- 20            # 67% extensive metabolizers
nPm <- 10            # 33% poor metabolizers
n   <- nEm + nPm

# Simulate per-subject parameters
clTrue <- c(
  exp(tclEm + rnorm(nEm, 0, sqrt(omegaCl))),
  exp(tclPm + rnorm(nPm, 0, sqrt(omegaCl)))
)
vTrue  <- exp(tv + rnorm(n, 0, sqrt(omegaV)))
kaTrue <- rep(exp(tka), n)
group  <- rep(c("EM (fast)", "PM (slow)"), c(nEm, nPm))

# Observation times
times <- c(0.5, 1, 2, 4, 6, 8, 12, 18, 24, 36, 48)

# One-compartment oral model
modSim <- rxode2({
  d/dt(depot)   <- -ka * depot
  d/dt(central) <- ka * depot - (cl / v) * central
  cp <- central / v
})

simRows <- vector("list", n)
for (i in seq_len(n)) {
  ev  <- et(amt = 100, time = 0) |> et(time = times)
  out <- rxSolve(modSim,
                 params = c(ka = kaTrue[i], cl = clTrue[i], v = vTrue[i]),
                 events = ev)
  dv  <- out$cp * exp(rnorm(length(times), 0, sigma))
  simRows[[i]] <- data.frame(
    ID    = i,
    time  = times,
    DV    = pmax(dv, 1e-6),
    group = group[i],
    AMT   = 0,
    EVID  = 0
  )
}

# Add dosing rows (EVID=1)
doseRows <- data.frame(
  ID = seq_len(n), time = 0, DV = NA_real_,
  group = group, AMT = 100, EVID = 1
)
simData <- rbind(doseRows, do.call(rbind, simRows))
simData <- simData[order(simData$ID, simData$time), ]
```

### Why a standard model fails

Plotting the raw concentration–time profiles immediately reveals the bimodality:

```{r plot-raw, fig.cap="Simulated PK profiles coloured by true metabolizer status. The 10-fold difference in clearance creates two non-overlapping trajectory bundles — a standard single-population model cannot capture this structure."}
obs <- simData[simData$EVID == 0, ]

# manual colour palette — no external packages needed
cols <- c("EM (fast)" = "#2166AC", "PM (slow)" = "#D6604D")

plot(NA, xlim = c(0, 48), ylim = c(0, max(obs$DV) * 1.05),
     xlab = "Time (hr)", ylab = "Concentration",
     main = "Observed profiles by true metabolizer status")
for (grp in c("EM (fast)", "PM (slow)")) {
  sub <- obs[obs$group == grp, ]
  for (id in unique(sub$ID)) {
    d <- sub[sub$ID == id, ]
    lines(d$time, d$DV, col = cols[grp], lwd = 1.2)
  }
}
legend("topright", legend = names(cols), col = cols, lwd = 2, bty = "n")
```

The two bundles of trajectories are separated by roughly an order of magnitude
in exposure — exactly the situation where a mixture model should be used.

## Model Specification

Mixture models are declared with `mix(expr1, prob1, expr2, ...)`:

- `expr1`, `expr2`, … are the parameter expressions for each component
- `prob1` is the prior mixing probability for the first component; the last
  component's probability is implicit (`1 - prob1`)
- One fewer probability parameter than components is required

```{r model}
twoPop <- function() {
  ini({
    tka   <- log(1.2)                   # log Ka
    tcl1  <- logit(0.8, 0.1, 200)      # log Cl, EM subpopulation
    tcl2  <- logit(0.8, 0.1, 200)      # log Cl, PM subpopulation
    tv    <- log(30)                     # log V
    p1    <- 0.67                        # prior prob of being an EM
    eta.cl ~ 0.09
    eta.v  ~ 0.04
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- mix(expit(tcl1 + eta.cl, 0.1, 200), p1, expit(tcl2 + eta.cl, 0.1, 200))
    v  <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
```

## Fitting

```{r fit}
fit := nlmixr2(twoPop, simData, "focei",
               control = foceiControl(print = 0))
print(fit)
```


```{r fit-numbers, echo=FALSE}
.bt <- function(f, n) signif(f$parFixedDf[n, "Back-transformed"], 3)
```

The back-transformed `p1` of **`r .bt(fit, "p1")`** is close to the true EM
prevalence (0.67). The two component clearances (`tcl1` →
`r .bt(fit, "tcl1")` L/hr, `tcl2` → `r .bt(fit, "tcl2")` L/hr) recover the
simulated 8 and 0.8 L/hr. `p1` now also carries a standard error and a
confidence interval — see
[Standard errors for the mixture proportions](#standard-errors-for-the-mixture-proportions)
for what they mean and how they are computed.

> The simulation above multiplies by `exp(rnorm(...))`, i.e. a proportional
> (log-normal) residual error, while the model fits an additive one
> (`add(add.sd)`) to keep the example short. The structural parameters and the
> mixture are recovered regardless, but do not read the estimated `add.sd`
> against the simulated `sigma` — they are not the same quantity.

### Why the estimates are good

A standard single-population FOCEI fit would instead produce a single clearance
estimate somewhere between 0.8 and 8 L/hr together with an implausibly large
BSV for CL (> 100%), because the model would try to explain the bimodal
exposure distribution through random effects alone.

## Accessing Mixture Results

After fitting, the subpopulation each subject was assigned to is available in
three places: two accessors on the fit object, and — for a model that asks for
it — as columns of the fitted table itself.

### `$mixNum` — best-fit subpopulation per subject

```{r mixnum}
head(fit$mixNum, 8)
```


`mixnum = 1` → EM (fast), `mixnum = 2` → PM (slow). In a real analysis you
would merge this back into the data frame and verify concordance with any
available genotype or phenotype data.

> **Two different things are called `mixnum`.** The `mixnum` *column of
> `$mixNum`*, above, is the component each subject was assigned to. The
> `mixnum` *reserved model variable* is the number of components in the model —
> a constant. The per-subject component in a model is `mixest`. This article
> names its output variables `component` and `nComponents` to keep the two
> apart.

### `$mixList` — per-component ETAs and posterior probabilities

```{r mixlist}
# ETAs and posterior probability of belonging to EM component
head(fit$mixList$mix1)
```


`prob` is the Bayesian posterior probability that subject $i$ belongs to
component $k$:

$$
P(k \mid i) \;\propto\; \exp\!\left(-\tfrac{1}{2}\,\mathrm{OBJI}(i,k)\right)
\times \pi_k
$$

where OBJI$(i,k)$ is the individual objective function for subject $i$ evaluated
under mixture component $k$ and $\pi_k$ is the estimated prior mixture
probability.

Subjects with `prob ≈ 1` are unambiguously assigned; subjects with intermediate
`prob` values (e.g. 0.3–0.7) are phenotypically ambiguous.

### Reading the component off the fit table

A model may also assign the reserved variables `mixest` (the component this
subject was fit under) and `mixnum` (how many components the model has) to
output variables, and they then appear as columns of the fitted table alongside
everything else the model computes:

```{r mixest-model}
twoPopComp <- function() {
  ini({
    tka   <- log(1.2)
    tcl1  <- logit(0.8, 0.1, 200)
    tcl2  <- logit(0.8, 0.1, 200)
    tv    <- log(30)
    p1    <- 0.67
    eta.cl ~ 0.09
    eta.v  ~ 0.04
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- mix(expit(tcl1 + eta.cl, 0.1, 200), p1, expit(tcl2 + eta.cl, 0.1, 200))
    v  <- exp(tv + eta.v)
    component   <- mixest   # which subpopulation this subject is in
    nComponents <- mixnum   # how many there are (a constant)
    linCmt() ~ add(add.sd)
  })
}

fitComp := nlmixr2(twoPopComp, simData, "focei",
                   control = foceiControl(print = 0))
```

```{r mixest-table}
head(as.data.frame(fitComp)[, c("ID", "TIME", "component", "nComponents",
                                "cl", "v")])
```


`component` is per subject and agrees with `$mixNum`; `nComponents` is the same
constant on every row. `IPRED` is the prediction under that subject's own
component. `PRED`, being the *population* prediction, takes one value per
component per time point, so a PRED-versus-time plot of a two-component mixture
shows two curves rather than one — that is expected, not a fitting problem. This is the most convenient route when you want the
component next to the individual parameters and predictions rather than in a
separate data frame — note that `cl` above is the clearance of the component the
subject was actually assigned to, not an average over components.

The name you assign to is yours; `mixest`/`mixnum` are the reserved variables on
the right-hand side. `mixunif`, the third reserved variable, is only meaningful
for *simulation* from a `mix()` model — in a fitted table it reports the
component that was supplied rather than a uniform draw, so use `mixest` there.

> These columns require an rxode2 carrying the fix described in the
> requirements note in the [Overview](#overview). On an older one they read `0`
> and the fit says so in `$runInfo`.

### Checking classification accuracy (when true labels are known)

In a simulation study the true labels can be compared directly:

```{r classification}
# Merge predicted mixture assignment with true labels
truth <- data.frame(ID = seq_len(n), trueGroup = group)
comp  <- merge(fit$mixNum, truth, by = "ID")
comp$predicted <- ifelse(comp$mixnum == 1, "EM (fast)", "PM (slow)")
table(Predicted = comp$predicted, True = comp$trueGroup)
```


Perfect classification in this example — the 10× clearance difference makes the
two populations easy to separate.

> **Check the label order before reading that table.** The `ifelse()` above
> hard-codes "component 1 is EM", which is only true because *this* fit happened
> to land that way. Nothing constrains which component gets which index, and a
> different seed, dataset or method can swap them. A swapped run makes the
> identical code report *zero*
> agreement, which looks like a failed fit and is not one. Confirm the order from
> the estimates — compare the back-transformed `tcl1`/`tcl2` against what you
> expect each subpopulation's clearance to be — before labelling the components.

## Interpreting the `$parFixed` Table

Mixture probability parameters (e.g. `p1`) are estimated on the mlogit scale
internally. `nlmixr2est` automatically back-transforms them to the probability
scale in the `Back-transformed` column. The back-transformation is applied
*jointly* across all mixture probability parameters via `rxode2::mexpit()`,
preserving the constraint that probabilities sum to 1 across components.

### Standard errors for the mixture proportions

`p1` gets a standard error, a %RSE and a confidence interval like any other
parameter. (Older versions of nlmixr2est reported `NA` here.) The quantity
behind them is the NONMEM 7 Technical Guide's own
construction, eq. (7.51)-(7.54): the mixture parameters' information matrix is
the outer product of the per-subject scores

$$
I_{a} \;=\; \sum_i g_{ia}\, g_{ia}^{\prime},
\qquad
g_{ia,\ell} \;=\; -2\,(r_{i\ell} - \pi_\ell)
$$

where $r_{i\ell}$ is subject $i$'s posterior responsibility for component
$\ell$ (the `prob` column of `$mixList`) and $\pi_\ell$ the estimated
proportion. That is the same score the estimation itself uses, so the SE costs
no extra model solves.

Three things are worth knowing about how it is reported.

**It is on the probability scale.** The proportions are *estimated* as a
multinomial logit, but reporting an SE there would put it next to an estimate
that is a probability. The covariance is therefore rotated with the full
Jacobian of `mexpit()`, $J = \mathrm{diag}(\pi) - \pi\pi^{\prime}$, so `$cov`,
the SE, the %RSE and the CI all sit on the same scale as the estimate. Using the
*full* Jacobian rather than just its diagonal is what carries the proportions'
correlations — with each other and with the structural thetas — onto that scale.

**The interval is a logit-scale interval.** A symmetric $\hat p \pm 1.96\,SE$
walks out of $(0,1)$ for a proportion near either end. The reported CI is
$\mathrm{expit}(\mathrm{logit}(\hat p) \pm z\,SE/(\hat p(1-\hat p)))$, which
stays inside $(0,1)$ and is asymmetric — which is the honest shape here.

**Near a boundary the SE shrinks toward zero.** The rotation carries a factor of
$\pi(1-\pi)$, so a proportion sitting at 0 or 1 gets a very small SE. That is
the correct delta-method answer but it reads as certainty, when really the Wald
approximation has stopped being informative; the fit says so in `$runInfo` when
it happens. Treat a near-boundary proportion as "this component may not be
supported by the data", not as a precise estimate.

Which covariance methods provide it:

| `covMethod` | Mixture proportion SE |
|---|---|
| `"r,s"`, `"r"`, `"s"` | yes (the default route for `focei` and its relatives) |
| `"imp"` | yes |
| `"analytic"` | declines for a mixture and falls back to the finite-difference sandwich, which does provide it |
| `linFim` / `fim` / `sa` (`saem`) | only when the fit is at the mixture's score-zero point; see below |

`saem` leaves the proportions out of the parameter vector its kernel converges,
so the block is appended from the fit's own responsibilities — and only when
$\pi_\ell = \overline{r}_{\cdot\ell}$ actually holds. An information matrix
describes the precision of a maximum-likelihood estimate, and away from that
point it would be a confident-looking number attached to something that is not
one. When the identity fails, the fit says so in `$runInfo` and reports no SE
rather than a misleading one.

For individual-level uncertainty — how sure the fit is about *this subject's*
membership, as opposed to the population proportion — use the `prob` columns of
`$mixList`.

## Using `$mixNum` in Downstream Analysis

```{r downstream}
library(dplyr)

fitData <- as.data.frame(fit)
fitAug  <- left_join(fitData, fit$mixNum, by = "ID")

fitAug |>
  group_by(mixnum) |>
  summarize(
    nSubjects   = n_distinct(ID),
    medianCl    = median(cl, na.rm = TRUE),
    medianIpred = median(IPRED, na.rm = TRUE),
    .groups = "drop"
  )
```


## Visualizing Posterior Probabilities

```{r plot-prob}
library(ggplot2)
library(dplyr)

probs <- bind_rows(
  mutate(fit$mixList$mix1, component = "Mix 1 (EM)"),
  mutate(fit$mixList$mix2, component = "Mix 2 (PM)")
)

ggplot(probs, aes(x = factor(ID), y = prob, fill = component)) +
  geom_col() +
  scale_fill_manual(values = c("Mix 1 (EM)" = "#2166AC", "Mix 2 (PM)" = "#D6604D")) +
  labs(x = "Subject ID", y = "Posterior probability",
       title  = "Mixture membership probabilities",
       fill   = NULL) +
  theme_bw() +
  theme(axis.text.x = element_text(size = 7))
```

Most subjects will show a bar that is essentially 100% one colour, confirming
clear separation of the two populations.


## Mixture Models with SAEM

SAEM accepts the same `mix()` model syntax as the gradient-based methods, so the
`twoPop` model defined above can be submitted to it directly:

```{r fit-saem}
fitSaem := nlmixr2(twoPop, simData, "saem",
                   control = saemControl(nBurn = 200, nEm = 300, print = 0))
print(fitSaem)
```

`tka` and `tv` get real, finite SEs; `tcl1`/`tcl2` (the mixture-owned fixed
effects) correctly show no SE rather than either an error or a fabricated
number — see the covariance callout below.  `p1` does now carry an SE: the
proportion is reported at its score-zero point, which is the condition the
mixture information matrix needs (see
[Standard errors for the mixture proportions](#standard-errors-for-the-mixture-proportions)).

Compare the back-transformed `p1` above against the true EM prevalence of
0.67, and check that `fitSaem$mixNum` is a real rather than a collapsed split:

```{r mixnum-saem-table}
table(fitSaem$mixNum$mixnum)
```

Cross-tabulated against the true labels.  Which component gets index 1 is not
fixed, so read the counts rather than the row labels (see the caution above):

```{r classification-saem}
compSaem <- merge(fitSaem$mixNum, truth, by = "ID")
compSaem$predicted <- ifelse(compSaem$mixnum == 1, "component 1", "component 2")
table(Predicted = compSaem$predicted, True = compSaem$trueGroup)
```

SAEM typically separates these two subpopulations less cleanly than FOCEi does
on the same data; the settings discussed next are what to reach for when it
does.

> **Two known limitations, both distinct from component separation itself.**
> First, `tcl1`/`tcl2` (the mixture-owned fixed effects) show no SE. The
> linearized Fisher-information calculation degrades gracefully now — `tka`
> and `tv` get real SEs instead of the whole covariance matrix failing — but
> `tcl1`/`tcl2` themselves remain genuinely under-identified from the
> linearization alone whenever few subjects are confidently assigned to the
> minority component, *whether or not* the components share an ETA (using
> split ETAs, see below, does not by itself fix this). Second, separation
> quality itself is sensitive to `nBurn`/`nEm` and to the
> `mixProbPriorN`/`mixProbMethod` settings described next — increasing
> `mixProbPriorN` to 30 and using `nBurn = 400, nEm = 800` moves it toward
> FOCEi's separation on this same simulated data. For
> a first look at a mixture model, or when precise per-subject classification
> matters, `focei` remains the more reliable default; SAEM is a usable
> alternative that may need more iterations and/or a higher `mixProbPriorN`
> to match it.

### Why SAEM needs help separating mixture components

The gradient-based methods use a **hard assignment**: every subject is solved
under each component, and the single best-fitting component is selected before
the sufficient statistics are updated — this optimizes each subject
independently on every iteration, so it cannot get stuck. SAEM instead uses a
**soft (probabilistic) assignment** that fits naturally into its
stochastic-approximation framework, but that same stochastic-approximation
mechanism makes the *mixing probability* itself prone to a runaway
positive-feedback collapse: the per-subject responsibility used to update it
is weighted by the *current* mixing probability, so a small, noisy drift
shrinks the responsibility of the shrinking component for every subject (not
just the ones that truly belong to the other component), which shrinks the
next iteration's estimate further. Left unchecked this can drive the mixing
probability all the way to exactly 0 or 1 well before the fixed effects have
separated — `saemControl()`'s `mixProbMethod` (below) exists specifically to
prevent that.

### Stabilizing the mixing-probability estimate: `mixProbMethod`

`saemControl()` exposes three ways to stabilize the mixing probability against
the collapse described above:

* **`mixProbMethod = "regress"` (the default).** Skip the per-iteration soft-EM
  responsibility step altogether: each subject is hard-classified to a
  component once, up front, and that membership is held fixed and fed into the
  solve as a regressor. This avoids the feedback loop by construction and is
  the lowest-bias option. The cost is variance rather than bias — an early
  misclassification is never revisited — so prefer `"regularized"` when
  membership is genuinely uncertain. Because membership is fixed, the `prob`
  column of `$mixList` is 0/1 under this option rather than a graded posterior;
  use one of the other two if you want soft weights.

* **`mixProbMethod = "regularized"`.** Blend `mixProbPriorN`
  pseudo-subjects, distributed according to the model's *initial* mixing
  probability (from `ini()`), into the responsibility average used at every
  iteration — equivalent to a Dirichlet-prior/MAP-EM correction on the mixing
  proportions. This changes what the estimate converges toward (pulled toward
  the initial guess, with the pull's relative strength shrinking as the
  number of subjects grows), in exchange for reliably preventing collapse.
  Larger `mixProbPriorN` and more `nBurn`/`nEm` iterations both improve
  separation quality, as shown above.

* **`mixProbMethod = "annealed"`.** Give the mixing-probability update its own
  step-size schedule that decays from iteration 1 (`mixProbStepExp`), instead
  of the shared burn-in step of 1 used for ordinary parameters. This only
  slows *how fast* the estimate can move, without biasing what it converges
  to — but for a systematic (rather than purely noise-driven) collapse,
  slowing the walk to the same wrong answer does not, by itself, prevent
  reaching it. Prefer this when you want to avoid any bias toward the initial
  mixing probability and believe the components are already reasonably
  separated.

```{r mixprob-method, eval=FALSE}
# more separation power, at the cost of more iterations and some bias
# toward the ini()-specified p1:
saemControl(mixProbMethod = "regularized", mixProbPriorN = 30,
            nBurn = 400, nEm = 800)
```

### How SAEM handles mixtures differently

The gradient-based methods use a **hard assignment**: every subject is solved
under each component, and the single best-fitting component is selected before
the sufficient statistics are updated. SAEM instead uses a **soft (probabilistic)
assignment** that fits naturally into its stochastic-approximation framework:

1. **Per-component MCMC.** At each iteration the SAEM kernel runs the MCMC
   sampling of the random effects *once per mixture component*, producing a
   separate set of posterior `phi` samples for each component.
2. **Posterior membership weights.** For every subject $i$ and component $k$ a
   posterior weight $a_{ik}$ is computed from the joint negative log-likelihood
   (observation loss **plus** the prior penalty on the random effects), combined
   with the current mixing probability $\pi_k$:
   $$
   a_{ik} \;=\; \frac{\pi_k \, \exp(-L_{ik})}{\sum_{j} \pi_j \, \exp(-L_{ij})}
   $$
   where $L_{ik}$ is the average joint negative log-likelihood of subject $i$
   under component $k$ across the MCMC samples. Using the *joint* loss (rather
   than the observation loss alone) is essential: the MCMC adapts each subject's
   ETAs to fit the data well under *every* component, so the prior penalty
   (the Mahalanobis distance of each sample from the component mean) is what
   discriminates which component truly fits.
3. **Weighted sufficient statistics.** The fixed-effect, $\Omega$, and residual
   sufficient statistics are accumulated as $a_{ik}$-weighted averages over all
   components rather than from a single selected component.
4. **Stochastic-approximation update of the mixing probability.** The population
   mixing probability is updated each iteration toward the mean posterior weight:
   $$
   \pi_k^{(\text{new})} \;=\; \pi_k \;+\; \gamma_{\text{iter}}\,
   \bigl(\bar a_{\cdot k} - \pi_k\bigr)
   $$
   with $\gamma_{\text{iter}}$ the usual SAEM step-size sequence.

Because the assignment is soft, SAEM does not discard information from the
non-selected components, which can improve stability when subpopulations are not
perfectly separated.

### Accessing SAEM mixture results

The same accessors are available as for the gradient-based fits, and return
values without error:

```{r saem-accessors}
fitSaem$mixNum         # most-likely component per subject (1-indexed)
fitSaem$mixList$mix1   # ID, ETAs, and posterior probability for component 1
fitSaem$mixProbabilities  # final population mixing probabilities (sum to 1)
```

A `mixest`/`mixnum` output variable works the same way here as for the
gradient-based methods — see
[Reading the component off the fit table](#reading-the-component-off-the-fit-table).
For SAEM the component reported there is the hard assignment `$mixNum` records,
which for `mixProbMethod = "regress"` (the default) is also what the fit itself
conditioned on.

For SAEM, the `prob` column in `$mixList` is the membership weight $a_{ik}$ the
SAEM engine computed directly (returned internally as the `mixWeights` matrix),
and `$mixNum` reports the component with the largest weight. **Under the default
`mixProbMethod = "regress"` that weight is 0/1**, because membership was
hard-classified once rather than updated as a responsibility — a valid MAP
assignment, but not a graded posterior. Switch to `"regularized"` or
`"annealed"` if you want to see how uncertain each subject's membership is.

`$etaExpected` holds the posterior-weighted ETAs used for shrinkage. These are
meaningful once `mixProbMethod` is keeping the fit stable, but their precision
(how many subjects land in the correct component) depends on the number of
iterations — and, under `"regularized"`, on `mixProbPriorN` — more so than for
the gradient-based methods.

### Shared variability across components (split ETAs)

A component expression may use its own random effect — for example
`mix(expit(tcl1 + eta.cl1, ...), p1, expit(tcl2 + eta.cl2, ...))`, where
`eta.cl1` and `eta.cl2` are separate ETAs sharing the same root name. SAEM
detects these *split ETAs* and collapses them back to a single between-subject
variability term in the reported `$omega`. The shared variance is computed with
the law of total variance using the posterior weights as component sizes:
$$
\operatorname{Var}(\eta) \;=\;
\underbrace{\sum_k w_k\,\sigma_k^2}_{\text{mean of variances}}
\;+\;
\underbrace{\sum_k w_k\,(\mu_k - \bar\mu)^2}_{\text{variance of means}},
\qquad w_k = \frac{\sum_i a_{ik}}{\sum_i \sum_j a_{ij}}
$$
so a single `eta.cl` term is reported even though each component was sampled with
its own random effect during estimation.

```{r fit-saem-split}
twoPopSplit <- function() {
  ini({
    tka   <- log(1.2)
    tcl1  <- logit(0.8, 0.1, 200)
    tcl2  <- logit(0.8, 0.1, 200)
    tv    <- log(30)
    p1    <- 0.67
    eta.cl1 ~ 0.09
    eta.cl2 ~ 0.09
    eta.v  ~ 0.04
    add.sd <- 0.2
  })
  model({
    ka <- exp(tka)
    cl <- mix(expit(tcl1 + eta.cl1, 0.1, 200), p1, expit(tcl2 + eta.cl2, 0.1, 200))
    v  <- exp(tv + eta.v)
    linCmt() ~ add(add.sd)
  })
}
fitSaemSplit := nlmixr2(twoPopSplit, simData, "saem",
                        control = saemControl(nBurn = 200, nEm = 300, print = 0,
                                              covMethod = "linFim"))
print(fitSaemSplit)
```

Split ETAs correctly separate the two clearances (5.98 and 0.71, near the true
8 and 0.8) instead of collapsing together — as with the non-split example
above, `tcl1`/`tcl2` still show no SE from the linearization, and this
particular run's reported `eta.cl` BSV (1.08 SD) is not trustworthy (true
value: 0.3 SD) — see the callout below.

#### Prefer a truly mu-referenced parameterization when you can

`tcl1 + eta.cl1` above is *nested inside* `expit(...)`, so `cl` itself is a
nonlinear function of the mu-referenced quantity. If your parameter doesn't
need a hard bound, write the mixture directly in terms of the mu-referenced
value with no wrapper, e.g. `cl <- mix(tcl1 + eta.cl1, p1, tcl2 + eta.cl2)`
with `tcl1`/`tcl2` initialized directly to a clearance guess (not a
log/logit-scale one). On the same simulated data this recovers **7.5–7.6**
and **0.64–0.67** across seeds — closer to the true 8/0.8 than the bounded
form above — because SAEM's per-component weighted update (used to keep
`tcl1`/`tcl2` from being diluted by each other, described next) is an exact
weighted-least-squares fit when the mu-referenced quantity has no nonlinear
wrapper, and a wrapped/bounded parameterization tends to compress the two
components closer together in the space the likelihood actually sees,
weakening separation. Prefer the bounded form only when you genuinely need a
hard constraint (e.g. a rate that must stay positive) that the data can't be
trusted to respect on its own.

> **No parameterization prevents label switching.** Whichever form you use,
> there is no way to pin down *which* mixture component ends up as component
> 1 vs. component 2 without adding a boundary/ordering constraint to the
> parameter — and adding one (e.g. wrapping in `logit()`/bounding `tcl1` below
> `tcl2`) requires a nonlinear transform, which is exactly the "wrapped" form
> the previous paragraph recommends against for separation quality. So the
> order of `tcl1`/`tcl2` (which one comes out larger) — and correspondingly
> which numeric label `mixNum` assigns to which true subpopulation — can
> differ from run to run or across seeds. Always check the *estimated
> values* (or cross-tabulate `mixNum` against known labels, as in the
> classification example above) rather than assuming `tcl1` is always "the
> fast/EM group."
>
> **The BSV/`$omega` issue above is independent of parameterization.**
> Whether you use the bounded or the mu-referenced form, split ETAs' reported
> variability can still come out wrong (too large or collapsed near zero) even
> when the fixed effects themselves separate correctly. Treat `$omega`/BSV%
> for split-ETA mixture components as unreliable for now; the point estimates
> of `tcl1`/`tcl2` and `p1` are the numbers to trust.

### An alternative sampling architecture: `mixSampleMethod = "msaem"`

The BSV problem above comes from *how* SAEM's default architecture explores
the random effects for a mixture. `saemControl(mixSampleMethod=)` selects
between two designs:

* **`"parallel"` (default).** Every iteration, one full MCMC chain is run
  *per component* for every subject (each chain explores as if that subject
  definitely belongs to that component), then the results are blended
  post hoc using posterior weights. This mirrors NONMEM's population
  mixture modeling (`$MIX`) and correctly estimates a single BSV shared
  across all components — but it has no way to obtain a clean per-component
  BSV for split-ETA models: each "wrong-hypothesis" chain still explores its
  non-owned column(s) as an unconstrained prior-only random walk, and that
  noise ends up in the blended statistics.

* **`"msaem"` (experimental).** Implements MSAEM (Lavielle & Mbogning 2014,
  *Statistics and Computing* 24(5)), the method Monolix uses for mixture
  models: only *one* random-effects trajectory is simulated per subject per
  iteration, and the discrete component label is analytically marginalized
  out via an exact posterior responsibility instead of being simulated —
  so there is nothing to blend after the fact. It also uses a model-aware
  stratified initialization (each subject is tested against every candidate
  component using the actual compiled model before fitting starts), which
  reliably separates split-ETA fixed effects even from the realistic
  symmetric starting point used throughout this vignette (`tcl1`/`tcl2`
  both initialized identically).

`"parallel"` runs `nMix` full chains per subject per iteration where
`"msaem"` runs one, so they are not compute-matched at equal `nmc` — set
`nmc` to roughly `nMix` times its default (`nmc = 6` for a 2-component
mixture here) for a fair comparison:

```{r fit-saem-msaem}
fitSaemMsaem := nlmixr2(twoPopSplit, simData, "saem",
                        control = saemControl(nBurn = 200, nEm = 300, nmc = 6,
                                              print = 0,
                                              mixSampleMethod = "msaem",
                                              covMethod = "linFim"))
print(fitSaemMsaem)
```

Fixed effects separate about as well as `"parallel"` — `tcl1` and `tcl2`
back-transform to `r .bt(fitSaemMsaem, "tcl1")` and
`r .bt(fitSaemMsaem, "tcl2")` against the true 0.8 and 8. The usual label-order
caveat applies: which component gets index 1 is not fixed, so compare the two
back-transformed clearances against what you expect each subpopulation to be
before reading anything else in the row.

The raw, unpooled per-component BSV (`fitSaemMsaem$env$saem0$Gamma2_phi1`) is
better than `"parallel"`'s, whose equivalent raw diagonal is *always* exactly
zero for split-ETA columns (the pooled `$omega` shown in `parFixed` only looks
reasonable because pooling adds in the between-component variance of the
means) — but it is still not a reliable recovery of the true 0.09.

And read `p1` against the component it actually belongs to. `p1` is the prior
probability of *component 1*, which in this fit is the component whose
clearance is `r .bt(fitSaemMsaem, "tcl1")` — so compare it to that
subpopulation's true fraction, not automatically to the 0.67 the earlier
examples use. That mismatch, not a fitting failure, is what makes a swapped run
look wrong at first glance.

**Bottom line:** `"msaem"`'s fixed-effect separation is as solid as
`"parallel"`'s, and its split-ETA BSV, while still not reliable, is
closer to the truth than `"parallel"`'s (which is always exactly zero
underneath the pooled report). It remains experimental — `mixProb` can be
less accurate, and BSV recovery is inconsistent across datasets/seeds.
Prefer `"parallel"` (the default) unless you are specifically evaluating
`"msaem"` or comparing the two.

## Other estimation families

Mixture support is not limited to the gradient-based methods and SAEM. Two more
families accept `mix()` and return the same `$mixNum`/`$mixList` accessors and
the same `mixest`/`mixnum` table columns.

### The EM / importance-sampling family: `imp`, `impmap`, `qrpem`

These are part of the same FOCEi family as the gradient-based methods and reach
the mixture the same way — ETAs are optimized for every subject x component
combination and the best-fitting component is selected, exactly as described in
[Accessing Mixture Results](#accessing-mixture-results) — so `$mixNum` and
`$mixList$prob` mean the same thing here as they do for `focei`. What differs is
how the individual marginal likelihood is obtained: by importance sampling
rather than by a Laplace-style linearization.

```{r imp-fit}
fitImp := nlmixr2(twoPop, simData, "imp", control = impControl(print = 0))
fitImp$parFixed
```

They are a reasonable middle ground: no linearization error, and, because the
component assignment is the family's hard assignment rather than a soft-EM
responsibility, no mixing-probability collapse to guard against — at the cost of
the Monte-Carlo sampling itself. `impmap` re-centres the sampling on the MAP
estimate and `qrpem` uses a quasi-random sequence, both of which reduce the
sampling variance relative to plain `imp` at the same number of draws.

### The nonparametric family: `npag`, `npb`

#### A nonparametric distribution is already a mixture

This needs stating plainly, because it is easy to reach for `mix()` here out of
habit and get nothing for it.

`npag` (adaptive grid) and `npb` (Dirichlet-process Bayesian) do not assume the
random effects are normal. They estimate a *discrete* distribution — a set of
support points $\phi_k$ with weights $w_k$ — and the marginal likelihood of a
subject is

$$
p(y_i) \;=\; \sum_k w_k \, p(y_i \mid \phi_k)
$$

which is a mixture already, over the support points, with as many components as
the data will support and each subject carrying a posterior over them. Two
distinct subpopulations do not need to be declared: the grid puts mass in two
places, and the two groups of subjects separate on their own. **If all you mean
is "clearance is bimodal", a nonparametric fit gives you that without `mix()`,
without committing to how many modes there are, and without needing initial
estimates for each one.**

#### So what does `mix()` add?

The two describe different things, and they sit at different levels of the
model:

| | What varies | How many | Estimated as |
|---|---|---|---|
| Nonparametric support | the **random effects** — where subjects sit in the parameter space | discovered from the data | support points and weights |
| `mix()` | the **structural model** — a different fixed-effect expression per subpopulation | declared by you | per-component thetas plus a proportion |

A support point is an unlabeled location in the random-effect space. A `mix()`
component is a claim that some subjects follow a *different equation*: a
different typical clearance parameter, a different absorption model, a
different anything you can write on the right-hand side. That is not something a
support distribution can express, because every support point is evaluated
through the same structural model.

So declare `mix()` on top of a nonparametric method when you want the things
only a declared mixture gives you — a named, estimated proportion you can report
and test, a per-component fixed effect with its own interpretation, a
per-subject component assignment tied to a mechanism (a genotype, a formulation,
a disease state) — and let the nonparametric distribution keep doing what it is
for, describing the spread *within* each of those components without assuming it
is normal. Reach for the nonparametric fit alone when the multimodality is the
finding rather than the hypothesis.

#### How the two are fitted together

One nonparametric distribution is shared across the declared components — the
support set is not re-estimated per component. Each subject is split into
per-component pseudo-subjects and the conditional likelihood is marginalized
over the components using the mixing proportions,
$p(y_i \mid \phi) = \sum_m \pi_m\, p(y_i \mid \phi, m)$. `npag` then updates
the proportion once per cycle with an EM step, holding the support points and
weights fixed; `npb` samples it inside its blocked Gibbs sweep — each subject
draws a component from its posterior responsibility and the proportions are
drawn from $\mathrm{Dirichlet}(1 + \text{component counts})$, with the
posterior-mean proportions reported in `$env$npbMixProb`. A `fix()`ed
proportion is held at its `ini()` value in both.

The components' own structural parameters (the per-component typical clearance,
say) are estimated alongside the residual parameters against that same
marginalized conditional likelihood — they are estimated, not held at their
initial values. Only the proportions are handled by the separate update step
above.

```{r np-fit}
fitNpag := nlmixr2(twoPop, simData, "npag", control = npagControl(print = 0))
fitNpag$parFixed
```

### Methods that decline a mixture

`emvi` and `fbvi` (mean-field variational inference) stop up front with

```
'twoPop' cannot have a mixture model (ie `mix()`) for the estimation routine 'emvi'
```

rather than silently fitting a model without the mixture structure.

`vae` **does** fit a mixture model. It follows the same
convention as the gradient-based methods -- every subject-component pair is
characterized by the encoder at its own eta, membership is the posterior
responsibility, and the proportions are estimated on the mlogit scale through
their analytic gradient -- so a `mix()` model can be submitted to it directly.
Earlier versions stopped during assembly rather than fitting.

## Choosing an Estimation Method

All six gradient-based methods handle mixture models using the same strategy:
ETAs are optimized separately for every subject × mixture combination and the
best-fitting component is selected. This works reliably in practice and gives
the cleanest separation. SAEM's scheme is described in
[Mixture Models with SAEM](#mixture-models-with-saem); its default,
`mixProbMethod = "regress"`, hard-classifies rather than collapsing, but it may
still separate components less precisely than the gradient-based methods on
small/weakly-separated datasets. Where membership is genuinely uncertain, switch
to `mixProbMethod = "regularized"` and raise `mixProbPriorN` and `nBurn`/`nEm`
(`mixProbPriorN` applies only to `"regularized"`; see that section for a worked
example):

| Method    | Interaction | Accuracy | Notes |
|-----------|:-----------:|:--------:|-------|
| `focei`   | Yes | High    | Recommended; full FOCE with interaction |
| `foce`    | No  | High    | Like FOCEI without η–ε interaction |
| `laplace` | Yes | High    | Good alternative to FOCEI |
| `agq`     | Yes | Highest | Most accurate; slowest |
| `fo`      | No  | Low     | Linearization; fast but less accurate |
| `foi`     | Yes | Low     | FO with interaction |
| `saem`    | Yes | Moderate–High | Hard-classifies by default (`mixProbMethod = "regress"`); for uncertain membership use `"regularized"` and tune `mixProbPriorN`/`nBurn`/`nEm`; see [Mixture Models with SAEM](#mixture-models-with-saem) |
| `imp` / `impmap` / `qrpem` | Yes | High | Same hard assignment as FOCEi, marginal likelihood by importance sampling; see [Other estimation families](#other-estimation-families) |
| `npag` / `npb` | Yes | — | Nonparametric random effects; marginalizes over components. Consider whether you need `mix()` at all — see [Other estimation families](#other-estimation-families) |

## Notes and Limitations

- **Computational cost** is multiplied by the number of mixture components,
  because the inner optimization runs for `nSub × nMix` individuals (or, for
  SAEM, `nMix` MCMC passes per iteration).
- **Identifiability** requires that the mixture components are well-separated
  in the data. More than 2–3 components is rarely supportable with typical
  PK/PD datasets.
- **A proportion sitting at 0 or 1** is the usual sign that a component is not
  supported by the data. Its reported SE will be very small, which is the
  correct delta-method answer and not evidence of precision — see
  [Standard errors for the mixture proportions](#standard-errors-for-the-mixture-proportions).
- The `mixest`, `mixnum`, and `mixunif` variables can be included in the model
  block for diagnostic purposes but do not affect the likelihood. `mixest` and
  `mixnum` report the fitted component and the component count in the fit table
  (see [Reading the component off the fit table](#reading-the-component-off-the-fit-table));
  `mixunif` is only meaningful when simulating.
- **Initial estimates matter**: provide lower bounds on `tcl1` and `tcl2` to
  prevent label-switching for focei during optimization. They matter at least
  as much for SAEM under its default `mixProbMethod = "regress"`, which
  hard-classifies every subject once, up front, and never revisits that
  assignment — so a poor start is not corrected by later iterations. Where
  membership is genuinely uncertain, prefer `"regularized"`.
- **SAEM** supports mixtures via an EM-responsibility (soft assignment) scheme
  whose default, `mixProbMethod = "regress"`, hard-classifies each subject once
  rather than updating responsibilities, and correctly separates split-ETA fixed
  effects under both `mixSampleMethod` options (see
  [Mixture Models with SAEM](#mixture-models-with-saem)); it typically needs
  more tuning than the gradient-based methods to match their separation
  precision — more `nBurn`/`nEm`, and, if you switch to `"regularized"` for
  uncertain membership, a larger `mixProbPriorN` (which applies to that option
  only). Two things it does not currently do
  well regardless of parameterization or `mixSampleMethod`: standard errors
  for the mixture-owned fixed effects (`tcl1`/`tcl2`-style parameters) come
  back as `NA` rather than a real SE (the mixture *proportion* does get one
  under the gradient-based methods — see
  [Standard errors for the mixture proportions](#standard-errors-for-the-mixture-proportions)),
  and `$omega`/BSV for split-ETA
  components can be unreliable (too large or collapsed near zero under
  `mixSampleMethod="parallel"`, the default; improved but still not
  consistently accurate under the experimental `"msaem"`) even when the
  fixed effects themselves separate correctly — trust the point estimates
  of `tcl1`/`tcl2`/`p1`, not their SEs or BSV%. Nested transforms inside
  `mix()` require **rxode2 >= 5.1.3** regardless of estimation method.
