---
title: "Simulations with individually-titrated dosing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    self_contained: no
vignette: >
  %\VignetteIndexEntry{Simulations with individually-titrated dosing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  out.width = "100%"
)
```

```{r setup}
library(nlmixr2)
library(ggplot2)
```

# The problem

Some simulations cannot be expressed as a dataset.  If the dose a
subject receives depends on what that same subject's simulated
concentration happened to be an hour earlier, then the event table
cannot be written down before the solve starts -- the dose is an
*output* of the simulation, not an input to it.

This is the ordinary shape of a titration trial:

- give a dose,
- come back at a protocol-defined visit,
- look at the subject's response,
- decide the next dose from what you saw.

`rxode2` (and therefore `nlmixr2`) can do this directly: the model
block may **push new events into the subject's own event history while
the model is being solved**.  This article works through a titration
example using those adaptive-dosing helpers.  The general reference for
the helpers themselves is the
[Adaptive Dosing with rxode2](https://nlmixr2.github.io/rxode2/articles/adaptive-dosing.html)
article; this one is about assembling them into a simulation you would
actually run.

All models below are written as `nlmixr2` **model functions** (`ini({})`
plus `model({})`), which is the form you would already be using for a
fit, and which lets the dosing helpers take their default arguments.

# A fixed-dose reference simulation

Start with an ordinary one-compartment model and a fixed 300 mg dose
every 24 hours.  Nothing adaptive is happening yet -- this is the
comparator.

```{r model-fixed}
oneCompartment <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv  <- log(31.5); label("V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <-  ka * depot - cl / v * center
    cp <- center / v
    cp ~ add(add.sd)
  })
}
```

The event table has the three protocol doses and hourly sampling for
ten subjects.  The seed is set immediately before the solve so that the
same ten sets of random effects can be re-drawn later and the two
simulations compared subject by subject:

```{r events-fixed}
evFixed <- et(amt = 300, cmt = "depot", time = c(0, 24, 48)) |>
  et(seq(0, 72, by = 1)) |>
  et(id = 1:10)

set.seed(42)
simFixed <- nlmixr2(oneCompartment, evFixed, est = "rxSolve")
```

```{r fig-fixed}
plot(simFixed, ipredSim) + xlab("Time (hr)") + ylab("Concentration")
```

# Titrating the dose inside the model

Now suppose the drug causes a dose-limiting adverse event, and that the
AE is driven by exposure: at each protocol visit (24 and 48 hours,
pre-dose) a subject with `cp > 1.5` is deemed to be having an AE and has
their next dose halved to 150 mg.  A subject below the threshold gets
the full 300 mg.

Two things make this work:

1. **`bolus()`** pushes the visit dose from inside `model({})`, so the
   dose amount can depend on `cp` at that moment.  Only the first dose
   -- the protocol-mandated 300 mg at time 0 -- stays in the event
   table.
2. **`hasAE` is a sticky variable.**  A left-hand-side variable that is
   not assigned at the current time keeps the value it had at the
   previous time rather than being recomputed, so the AE status
   determined at a visit carries forward between visits.  It starts out
   `NA`, which is how you detect the first record for a subject and
   initialize it.  (Sticky variables are covered in detail in the
   [rxode2 sticky variables](https://nlmixr2.github.io/rxode2/articles/rxode2-sticky.html)
   article.)

```{r model-titrated}
oneCompartmentTitrated <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv  <- log(31.5); label("V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <-  ka * depot - cl / v * center
    cp <- center / v

    # Protocol visits: 24 and 48 hours.  The `t > 0` guard keeps the
    # rule from firing on the time-zero dose, which is protocol-fixed.
    isVisit <- (t > 0 && t <= 48 && t %% 24 == 0)

    # `hasAE` is NA on the first record of each subject; after that it
    # is sticky and keeps whatever the last visit assigned.
    if (is.na(hasAE)) {
      hasAE <- 0
    } else if (isVisit) {
      hasAE <- cp > 1.5
    }

    # A subject having an AE gets half a dose.
    if (isVisit) {
      bolus(300 * (1 - 0.5 * hasAE), depot)
    }

    cp ~ add(add.sd)
  })
}
```

The event table now contains only the time-zero dose and the sampling
grid.  The 24- and 48-hour doses do not appear anywhere in the data --
the model creates them:

```{r events-titrated}
evTitrated <- et(amt = 300, cmt = "depot", time = 0) |>
  et(seq(0, 72, by = 1)) |>
  et(id = 1:10)

set.seed(42)
simTitrated <- nlmixr2(oneCompartmentTitrated, evTitrated, est = "rxSolve")
```

`hasAE` is returned with the simulation, so the dosing decision is
auditable after the fact:

```{r show-decisions}
simTitratedDf <- as.data.frame(simTitrated)
subset(simTitratedDf, time %in% c(24, 48),
       select = c(id, time, cp, hasAE, depot))
```

Read the rows at the visits: `cp` is the pre-dose concentration that
drove the decision, `hasAE` is the decision, and `depot` shows the dose
that was actually pushed -- 300 or 150, plus whatever little was left
in the depot from the previous dose.  Subjects 7 and 8 are the
interesting ones -- their status changes between the two visits, in
opposite directions, so they are dosed differently at 24 and 48 hours.

```{r fig-titrated}
ggplot(simTitratedDf,
       aes(x = time, y = ipredSim, group = id,
           col = factor(hasAE, levels = c(0, 1), labels = c("No", "Yes")))) +
  geom_line(alpha = 0.5) +
  geom_hline(yintercept = 1.5, linetype = "dashed") +
  labs(x = "Time (hr)", y = "Concentration", colour = "Has AE?")
```

The dashed line is the AE threshold and subjects are coloured by their
current AE status, so a curve changes colour at a visit where the status
changed.

Because both simulations were seeded identically and the models share
the same random-effect structure, the same ten subjects appear in each
and the effect of the titration rule can be read off directly:

```{r fig-compare}
simCompare <- rbind(
  transform(as.data.frame(simFixed)[, c("id", "time", "ipredSim")],
            regimen = "Fixed 300 mg"),
  transform(simTitratedDf[, c("id", "time", "ipredSim")],
            regimen = "Titrated")
)

ggplot(simCompare, aes(x = time, y = ipredSim, colour = regimen)) +
  geom_line() +
  facet_wrap(~ id, ncol = 5) +
  labs(x = "Time (hr)", y = "Concentration", colour = NULL) +
  theme(legend.position = "bottom")
```

Subjects whose exposure never crossed the threshold (1, 3, 6 and 10)
get identical curves under both regimens -- the rule simply never
fired, so only one curve is visible in those panels.  The high-exposure
subjects (2 and 5 most clearly) separate after 24 hours, which is
exactly the intent of the titration.

# Variations

Because the decision is ordinary R-like model code, the rule can be
anything you can write down.  Three common variations:

## Dose holds and multi-level titration

`bolus()` only has to be called when you actually want a dose.  Skipping
the call is a dose hold, and `else if` chains give you a titration
ladder:

```{r model-ladder}
oneCompartmentLadder <- function() {
  ini({
    tka <- log(1.57); label("Ka")
    tcl <- log(2.72); label("Cl")
    tv  <- log(31.5); label("V")
    eta.ka ~ 0.6
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)
    d/dt(depot)  <- -ka * depot
    d/dt(center) <-  ka * depot - cl / v * center
    cp <- center / v

    isVisit <- (t > 0 && t <= 48 && t %% 24 == 0)

    if (is.na(doseLevel)) {
      doseLevel <- 300
    } else if (isVisit) {
      if (cp > 3) {
        doseLevel <- 0      # hold: too much exposure
      } else if (cp > 1.5) {
        doseLevel <- 150    # reduce
      } else {
        doseLevel <- 300    # full dose
      }
    }

    if (isVisit && doseLevel > 0) {
      bolus(doseLevel, depot)
    }

    cp ~ add(add.sd)
  })
}

set.seed(42)
simLadder <- nlmixr2(oneCompartmentLadder, evTitrated, est = "rxSolve")

subset(as.data.frame(simLadder), time %in% c(24, 48),
       select = c(id, time, cp, doseLevel))
```

Note the `doseLevel > 0` guard on the `bolus()` call: a held dose is a
dose that is never pushed, not a dose of zero.

## Carrying a decision forward permanently

The rule above re-assesses at every visit, so a subject can recover to
full dose.  If instead an AE should permanently commit a subject to the
reduced dose, only ever assign in one direction and let stickiness do
the rest:

```{r snippet-persist, eval=FALSE}
if (is.na(hasAE)) {
  hasAE <- 0
} else if (isVisit && cp > 1.5) {
  hasAE <- 1   # once set, never reset -- stickiness carries it forward
}
```

Because unassigned sticky variables keep their previous value, no
`hasAE <- hasAE` self-assignment is needed to make the value persist.

## Titrating an infusion instead of a bolus

`infuse()` and `infuseDur()` push infusions rather than boluses.  They
differ in what is held fixed when the delivered amount changes:

- `infuse(amt, rate, cmt)` fixes the **rate**, so the duration follows
  from `amt / rate`.  Halving the amount halves the infusion time.
- `infuseDur(amt, dur, cmt)` fixes the **duration**, so the rate is
  computed from the amount actually delivered.  Halving the amount
  halves the rate over the same window.

That distinction matters when bioavailability `f()` is also in play,
since `f()` changes the realized amount.

```{r model-infusion}
oneCompartmentInfusion <- function() {
  ini({
    tcl <- log(2.72); label("Cl")
    tv  <- log(31.5); label("V")
    eta.cl ~ 0.3
    eta.v  ~ 0.1
    add.sd <- 0.7
  })
  model({
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv + eta.v)
    d/dt(center) <- -cl / v * center
    cp <- center / v

    isVisit <- (t > 0 && t <= 48 && t %% 24 == 0)

    if (is.na(hasAE)) {
      hasAE <- 0
    } else if (isVisit) {
      hasAE <- cp > 1.5
    }

    # 2 hour infusion at each visit; a reduced amount is given over a
    # correspondingly shorter time because the pump rate is fixed.
    if (isVisit) {
      infuse(amt = 300 * (1 - 0.5 * hasAE), rate = 150, cmt = center)
    }

    cp ~ add(add.sd)
  })
}

evInfusion <- et(amt = 300, rate = 150, cmt = "center", time = 0) |>
  et(seq(0, 72, by = 0.5)) |>
  et(id = 1:10)

set.seed(42)
simInfusion <- nlmixr2(oneCompartmentInfusion, evInfusion, est = "rxSolve")
```

```{r fig-infusion}
plot(simInfusion, ipredSim) + xlab("Time (hr)") + ylab("Concentration")
```

# Practical notes

### The solver has to visit the decision time

Adaptive logic only runs when the solver is actually at a time point.
If the visit is at 24 hours and 24 hours is not in the event table, the
rule never fires and the dose is silently never given.  In this article
the hourly sampling grid `et(seq(0, 72, by = 1))` covers the visits, so
`t %% 24 == 0` is enough.

If you have a sparse sampling design, `mtime()` will make the solver
stop at a time that is not otherwise in the event table:

```{r snippet-mtime, eval=FALSE}
mtime(visit1) <- 24
if (t == visit1) {
  bolus(300, depot)
}
```

**But do not combine the two.** In a model function, if `mtime()` names
a time that is *already* in the event table, the dose is pushed twice
([rxode2#1152](https://github.com/nlmixr2/rxode2/issues/1152)).  Use
`mtime()` only for decision times the sampling grid does not already
cover, and a comparison guard like `isVisit` above when it does.

This one is worth checking rather than trusting, because it is invisible
in the solved concentrations -- the doubling only shows up in the state
at the *next* time point.  `addDosing = TRUE` reports the pushed events
directly:

```{r snippet-adddosing, eval=FALSE}
sim <- rxSolve(oneCompartmentTitrated, evTitrated, addDosing = TRUE)
subset(as.data.frame(sim), !is.na(evid) & evid != 0,
       select = c(id, time, evid, amt))
```

### Guard against repeated firing

A bare condition like `if (cp > 0.5)` stays true over many consecutive
time points, so it will push a dose at every one of them.  Anchor the
rule to a visit window -- that is what `isVisit` is doing in every model
above.

### `maxExtra`

Every pushed event counts against the `maxExtra` limit in `rxSolve()`.
If a rule accidentally creates a cascade, `rxode2` stops with a
`maxExtra` error rather than growing the event history without bound.
A `maxExtra` error usually means the guard on a condition is too loose,
not that the limit is too small.

### Check the pushed doses

The decision variables (`hasAE`, `doseLevel`) come back in the
simulated data frame, and the pushed amounts show up in the
compartment.  Both are worth inspecting before trusting a titration
simulation -- an off-by-one in a guard changes the answer without
producing an error.

# Summary

- Dose decisions that depend on the simulated response belong in
  `model({})`, not in the dataset.
- `bolus()`, `infuse()`, and `infuseDur()` push those doses at solve
  time; skipping the call is a dose hold.
- Sticky variables (`NA` until first assigned, then carried forward)
  hold the decision between visits with no self-assignment needed.
- Guard the rule so it fires only at protocol decision times, and make
  sure the solver visits them exactly once.

---

*This article grew out of
[nlmixr2#173](https://github.com/nlmixr2/nlmixr2/issues/173) and
@billdenney's draft in
[nlmixr2#174](https://github.com/nlmixr2/nlmixr2/pull/174), which asked
how to make a titration decision persist across time points.  The
answer changed since that draft: the `hasAE <- hasAE` self-assignment is
no longer needed because model variables are sticky, and the dose
adjustment itself no longer has to be smuggled in through
bioavailability -- `bolus()` can push the adjusted dose directly.
@billdenney, does this cover what you were after in #174?*
